---
title: "Catchment Delineation Workflow"
subtitle: "A practical guide to the stream and lake delineation pipelines"
author: "Sam Woodman — Natural Resources Canada"
date: today
format:
  html:
    toc: true
    toc-depth: 3
    toc-location: left
    toc-title: "Contents"
    code-fold: true
    code-tools: true
    code-link: true
    theme: flatly
    highlight-style: github
    embed-resources: true
    smooth-scroll: true
    number-sections: true
    fig-width: 7.5
    fig-height: 5.5
  pdf:
    toc: true
    toc-depth: 3
    number-sections: true
    documentclass: article
    fig-pos: "H"
    fig-width: 6.5
    fig-height: 4.8
    geometry:
      - margin=2.5cm
execute:
  echo: true
  warning: false
  message: false
  cache: true
---

```{r}
#| label: setup
#| include: false
#| cache: false

library(sf)
library(terra)
library(dplyr)
library(tidyr)
library(purrr)
library(stringr)
library(forcats)
library(readr)
library(tibble)
library(ggplot2)
library(tidyterra)
library(gt)
library(scales)
library(patchwork)

# This document is read-only with respect to the workflow: every figure and
# table below loads files that already exist under cache/ and output/ from a
# completed run of workflow/CELESTE/run_celeste.R and workflow/run_cam_lakes.R.
# No delineation, conditioning, or hydroweight step is re-run to build it, so
# it renders in well under a minute rather than hours. Pipeline code shown
# alongside each stage (grey, non-executed blocks) is reference — copy it into
# an R session if you actually need to run that step.
#
# NOTE: the stream worked example below reads output/CELESTE/ — CELESTE's
# pre-engine run, kept on disk deliberately (not deleted; see git history for
# the retired non-modular run_celeste.R that produced it). The CURRENT
# run_celeste.R runs on the modular engine and writes to
# output/CELESTE_engine/ instead. The pipeline SHAPE shown below (group
# manifest -> terrain -> delineate -> hydroweight) is unchanged either way;
# only the file paths differ if you want to point this guide at a fresh
# engine run instead.
#
# autodep = TRUE lets chunk caching (execute: cache: true, above) correctly
# invalidate a cached chunk when an *upstream* chunk it reads variables from
# changes, even though this chunk itself didn't.
knitr::opts_chunk$set(fig.align = "center", autodep = TRUE)

# -- Shared plotting helpers --------------------------------------------------

# Fixed-order, colourblind-safe base palette (Okabe & Ito, 2008). Named
# palettes are derived from it below and used (never re-cycled to a
# different meaning) so a given site/lake/direction keeps the same colour
# wherever it appears in the document — including across figures where a
# factor's sort order differs, since named lookup is used, not position.
pal_okabe_ito <- c(
  "#E69F00", "#56B4E9", "#009E73", "#F0E442",
  "#0072B2", "#D55E00", "#CC79A7", "#000000"
)

pal_nbe <- setNames(pal_okabe_ito, paste0("NBE", 1:8))

# Hand-picked (not positional) so none of the 4 are confusable — Okabe-Ito's
# two blues in particular are avoided here.
pal_lakes <- c(
  Clearwater = "#E69F00",
  Swan       = "#009E73",
  Tilton     = "#D55E00",
  Lohi       = "#CC79A7"
)

# WhiteboxTools' D8 pointer uses the ESRI encoding: 1=E, 2=SE, 4=S, 8=SW,
# 16=W, 32=NW, 64=N, 128=NE.
pal_dir <- setNames(pal_okabe_ito, c("E", "SE", "S", "SW", "W", "NW", "N", "NE"))

theme_cw <- function(base_size = 11) {
  theme_minimal(base_size = base_size) +
    theme(
      legend.position   = "bottom",
      panel.grid.minor  = element_blank(),
      plot.title        = element_text(face = "bold", size = rel(1.05)),
      plot.subtitle      = element_text(colour = "grey35"),
      strip.text         = element_text(face = "bold")
    )
}

# A hillshade + elevation "relief" figure is used repeatedly (group AOI,
# per-site DEM, per-lake DEM). Centralising it keeps every occurrence
# consistent and keeps individual chunks short.
plot_relief <- function(hillshade, dem = NULL, title = NULL, subtitle = NULL) {
  p <- ggplot() +
    tidyterra::geom_spatraster(data = hillshade, show.legend = FALSE) +
    scale_fill_gradient(low = "grey5", high = "grey95", na.value = NA)
  if (!is.null(dem)) {
    p <- p +
      ggnewscale::new_scale_fill() +
      tidyterra::geom_spatraster(data = dem, alpha = 0.55) +
      scale_fill_viridis_c(option = "viridis", name = "Elevation (m)", na.value = NA)
  }
  p + coord_sf(expand = FALSE) + theme_cw() +
    labs(title = title, subtitle = subtitle, x = NULL, y = NULL)
}

is_html <- knitr::is_html_output()
```

::: {.callout-note}
This guide was drafted with assistance from Claude (Anthropic), reading directly from this repository's code and cached outputs.
:::

## Overview

This workflow delineates catchments from pour-point coordinates for sites across Canada and computes morphometric and distance-weighted landscape metrics for each one. It is **project-agnostic**: a sites table and a handful of shared external datasets go in, self-contained per-site output folders come out.

Two pipelines currently run on it, sharing most of their machinery but differing wherever the feature type forces a difference — what the "outflow" target is, and whether flow products are prepared per-group or once for the whole project. This guide uses a real, already-processed example of each:

| | **Stream pipeline** | **Lake pipeline** |
|---|---|---|
| Feature type | River/stream pour point | Lake polygon |
| Example used below | `NBE` group — 8 New Brunswick sites (`NBE1`–`NBE8`) | `Clearwater`, `Swan`, `Tilton`, `Lohi` |
| Runner | `workflow/CELESTE/run_celeste.R` (engine-based; this example shows the retained pre-engine run) | `workflow/run_cam_lakes.R` |
| CRS | EPSG:3979 (Canada Atlas Lambert) | EPSG:3161 (Ontario MNR Lambert) |
| DEM source | MRDEM 30 m national VRT | OIH Enforced DEM (Ontario) |
| Hydroweight `target_O` | Snapped pour **point** (no area) | Lake **polygon** (has area → `remove_region`) |
| DEM/flow products | Per-`group_id` (HydroBasins-derived AOI) | One project-wide raster |
| LOI(s) demonstrated | NDVI, NDVI trend, harvest/regen | CANLCC land cover |

### Two-tier processing: groups and sites

Expensive DEM conditioning (breach depressions, flow direction, flow accumulation) is run once per **group** — a hydrologically coherent region sharing a DEM — never once per site. A **site** is a single pour point (or lake); several sites sharing a `group_id` reuse the same conditioned DEM and flow-accumulation raster. In the stream pipeline, `NBE` is one such group: all 8 NBE sites share `cache/NBE/dem.tif`, `flow_pointer.tif`, and `flow_accum.tif`.

::: {.columns}

::: {.column width="47%"}
**Stream pipeline**

```
Sites table
   │
   ▼
1. Group manifest
2. Prepare DEM
3. Burn NHN streams (optional)
4. WhiteboxTools conditioning
5. Delineate sites
6. Remove upstream (nesting)
7. Reclip outputs
8. Catchment metrics
9. Hydroweight LOIs
```
:::

::: {.column width="6%"}
:::

::: {.column width="47%"}
**Lake pipeline**

```
Lake sites table
   │
   ▼
1. Match lake polygons
2. Prepare DEM/flow
3. Delineate lakes
4. Remove upstream lakes
5. Reclip outputs
6. Catchment metrics
7. Hydroweight LOIs
```
:::

:::

### Repository layout

```
workflow/
├── R/
│   ├── utils.R                   shared: logging, validation, AOI helpers
│   ├── stream/                   reusable stream/river building blocks —
│   │   │                         used directly by engine/, not a pipeline
│   │   │                         of their own (old numbered-stage files
│   │   │                         retired; see git history). No numeric
│   │   │                         prefixes — order is set by the caller.
│   │   ├── group_sites.R
│   │   ├── burn_streams.R
│   │   ├── delineate_sites.R
│   │   ├── 99_rerun_sites        pre-engine only; kept for an unrelated script
│   │   └── hydroweight_attributes.R
│   ├── lake/                     reusable lake building blocks, same role
│   │   │                         as stream/ (old 02_prepare_oih_dem.R /
│   │   │                         03_delineate_lakes.R retired; terrain
│   │   │                         prep + delineation now go through engine/)
│   │   ├── match_lake_polygons.R
│   │   ├── remove_upstream_lakes.R
│   │   └── hydroweight_attributes.R
│   ├── engine/                   modular, input-driven delineation engine —
│   │   │                         standard approach for every project now
│   │   ├── 00_resolve_config.R
│   │   ├── 01_build_group_manifest.R
│   │   ├── 02_prepare_terrain.R
│   │   ├── 03_prepare_streams_burn.R
│   │   ├── 04_delineate_site.R
│   │   └── 99_rerun_sites
│   ├── remove_upstream.R         shared (stream nesting)
│   ├── reclip_outputs.R          shared (stream + lake)
│   └── catchment_metrics.R       shared (stream + lake)
├── raster_attributes.R           generic LOI-prep toolbox (project-agnostic)
├── CELESTE/                      CELESTE runner (engine-based) + LOI prep
│   ├── run_celeste.R
│   ├── prepare_ndvi.R
│   ├── prepare_ndvi_trend.R
│   ├── prepare_harvest_regen.R
│   └── tidy_outputs.R            long-format plotting-ready tables
├── CAM/                          CAM streams runner (engine-based) + LOI prep
│   ├── run_cam_streams.R
│   ├── prepare_ndvi.R
│   ├── prepare_ndvi_trend.R
│   ├── prepare_harvest_regen.R
│   ├── upload_catchments_to_gee.R
│   └── tidy_outputs.R
├── templates/
│   └── run_engine_template.R     blank TODO-annotated skeleton for a new project
└── run_cam_lakes.R                CAM lakes runner (engine-based)

code/
└── reset_workflow.R              cache/output reset helpers

cache/<group_id>/  or  cache/CAM/       group-level rasters (auto-created)
output/<site_id>/  or  output/CAM/…     per-site outputs (auto-created)
```

Anything specific to one pipeline — the runner and its LOI `prepare_*.R` scripts — lives in its own folder (`workflow/CELESTE/` for the stream pipeline; the lake pipeline doesn't need one yet, since `run_cam_lakes.R` is still self-contained). This keeps one pipeline's data-prep quirks from ever needing to be reconciled with the other's.

---

# Worked example A — Stream delineation (`NBE` group)

`NBE1`–`NBE8` are 8 New Brunswick pour points sharing `group_id = "NBE"`. Every figure in this section reads directly from `cache/NBE/` and `output/CELESTE/NBE*/` — files already produced by CELESTE's retained pre-engine run (see the note at the top of this document; the current `run_celeste.R` is engine-based and writes to `output/CELESTE_engine/` instead — the pipeline shape below is unchanged either way).

## Stage 1 — Group manifest

`build_group_manifest()` uses HydroBasins to build one processing AOI per `group_id`, unioning the basins containing that group's sites. All 9 stages downstream for a given group operate inside this one AOI.

```{r}
#| label: nbe-group-map
#| fig-cap: "NBE pour points on an OpenStreetMap basemap, zoomed to the site cluster. Inset: the full group AOI within New Brunswick."
group_aois <- st_read("cache/group_aois.gpkg", quiet = TRUE)
nbe_aoi <- group_aois |> dplyr::filter(group_id == "NBE")

nbe_pour_points <- list.files("output/CELESTE", pattern = "^NBE[0-9]$", full.names = TRUE) |>
  map(\(d) st_read(file.path(d, "pour_point.gpkg"), quiet = TRUE)) |>
  bind_rows()

# NBE1's catchment is reused as "the example catchment" by several later
# stages, so its extent stays consistent throughout this section.
nbe1_catchment <- st_read("output/CELESTE/NBE1/catchment.gpkg", quiet = TRUE)

zoom_bbox <- st_bbox(st_buffer(nbe_pour_points, 3000))

# maptiles::get_tiles() reprojects into whatever CRS its input sfc uses (here
# EPSG:3979), so the tile raster and coord_sf() stay in the same CRS — plain
# ggspatial::annotation_map_tile() assumes lon/lat and silently mis-locates
# non-4326 data, confirmed while building this figure.
basemap <- maptiles::get_tiles(st_as_sfc(zoom_bbox), provider = "OpenStreetMap", crop = TRUE, zoom = 12)

p_main <- ggplot() +
  tidyterra::geom_spatraster_rgb(data = basemap) +
  geom_sf(data = nbe_aoi, fill = NA, colour = "grey20", linewidth = 0.6) +
  geom_sf(data = nbe_pour_points, aes(colour = site_id), size = 2.6) +
  coord_sf(xlim = c(zoom_bbox["xmin"], zoom_bbox["xmax"]),
           ylim = c(zoom_bbox["ymin"], zoom_bbox["ymax"]), expand = FALSE) +
  scale_colour_manual(values = pal_nbe, name = "Site") +
  theme_cw() +
  labs(title = "NBE pour points",
       subtitle = glue::glue("buffer_m = {nbe_aoi$buffer_m}, burn_streams = {nbe_aoi$burn_streams}, n_sites = {nbe_aoi$n_sites}"))

nb <- rnaturalearth::ne_states(country = "Canada", returnclass = "sf") |>
  dplyr::filter(name == "New Brunswick") |>
  st_transform(3979)

p_inset <- ggplot() +
  geom_sf(data = nb, fill = "grey95", colour = "grey50") +
  geom_sf(data = nbe_aoi, fill = "#D55E00", colour = NA, alpha = 0.85) +
  theme_void() +
  theme(panel.background = element_rect(fill = "white", colour = "grey60"))

p_main + inset_element(p_inset, left = 0.01, bottom = 0.6, right = 0.4, top = 0.99)
```

```{r}
#| label: group-manifest-code
#| eval: false
#| code-fold: false
group_manifest <- build_group_manifest(
  sites            = sites,
  output_dir       = here("output/CELESTE"),
  cache_dir        = here("cache"),
  hydrobasins_dir  = HYDROBASINS_DIR,
  default_buffer_m = 1000
)
```

::: {.callout-warning}
Always inspect group AOIs (`cache/group_aois.gpkg`) before Stage 2. A group AOI that's too small truncates upstream area and produces an incorrect catchment — the fix is to raise `aoi_buffer_m` for that group, clear its cache, and rerun from Stage 2.
:::

## Stage 2 — Prepare DEM

The MRDEM is a GDAL VRT streamed from cloud tiles — `prepare_dem()` crops it to each group AOI and writes `cache/<group_id>/dem.tif`, a rectangular crop (never masked to the AOI polygon — masking would introduce NoData at the boundary that WhiteboxTools reads as depressions, corrupting flow direction near the edge). `cache/NBE/dem.tif` itself is ~100M cells (the full group AOI shown in the Stage 1 inset); the figure below crops it down to the example catchment's extent for display.

```{r}
#| label: nbe-dem
#| fig-cap: "Group DEM cropped to the NBE1 catchment extent, for display — prepare_dem() itself crops to the full group AOI, not this window."
dem_window   <- st_buffer(st_as_sfc(st_bbox(nbe1_catchment)), 1500)
nbe_dem_win  <- rast("cache/NBE/dem.tif") |> terra::crop(vect(dem_window))
nbe_hill_win <- terra::shade(
  terra::terrain(nbe_dem_win, "slope", unit = "radians"),
  terra::terrain(nbe_dem_win, "aspect", unit = "radians")
)

plot_relief(nbe_hill_win, nbe_dem_win, title = "Group DEM near NBE1") +
  geom_sf(data = nbe1_catchment, fill = NA, colour = "white", linewidth = 0.7)
```

```{r}
#| label: prepare-dem-code
#| eval: false
#| code-fold: false
prepare_dem(group_manifest, mrdem_vrt = MRDEM_VRT_PATH, target_crs = 3979)
```

## Stage 3 — Burn NHN streams (optional, per group)

When `burn_streams = TRUE` for a group, NHN flowlines are merged (fixing two known geometry issues — dropping M-coordinates and casting to a single `MULTILINESTRING` type) and burned into the DEM before conditioning, anchoring flow paths to mapped channels. NBE has `burn_streams = TRUE`; its cached NHN layers cover the whole group AOI (59,701 flowline features), so the figure below crops to a small window around `NBE1` rather than plotting the full group.

```{r}
#| label: nbe-streams
#| fig-cap: "NHN flowlines and waterbodies burned into the DEM near NBE1 (small window; the full NBE cache holds 59,701 flowline features across the group AOI)."
window    <- st_buffer(st_as_sfc(st_bbox(nbe1_catchment)), 2000)
window_bb <- st_bbox(window)

flowlines_win   <- st_read("cache/NBE/flowlines.gpkg", quiet = TRUE, wkt_filter = st_as_text(window))
waterbodies_win <- st_read("cache/NBE/waterbodies.gpkg", quiet = TRUE, wkt_filter = st_as_text(window))

# wkt_filter keeps any feature that *intersects* the window, not just the
# part of it inside the window — a flowline crossing the edge can extend far
# beyond it, so coord_sf() must set the crop explicitly rather than let it
# auto-fit to every layer's full (unclipped) extent.
ggplot() +
  geom_sf(data = window, fill = "grey97", colour = NA) +
  geom_sf(data = waterbodies_win, fill = "#a6cee3", colour = NA) +
  geom_sf(data = flowlines_win, colour = "#1f78b4", linewidth = 0.4) +
  geom_sf(data = nbe1_catchment, fill = NA, colour = "#D55E00", linewidth = 0.6) +
  coord_sf(xlim = c(window_bb["xmin"], window_bb["xmax"]),
           ylim = c(window_bb["ymin"], window_bb["ymax"]), expand = FALSE) +
  theme_cw() +
  labs(title = "NHN network near NBE1", subtitle = "Orange = NBE1 unclipped catchment boundary")
```

```{r}
#| label: nhn-geom-fix
#| eval: false
#| code-fold: false
# Applied internally in burn_streams_into_dem() before WhiteboxTools sees it:
flowlines |>
  sf::st_zm(drop = TRUE, what = "ZM") |>
  sf::st_cast("MULTILINESTRING") |>
  sf::st_write(streams_shp, delete_dsn = TRUE, quiet = TRUE)
```

::: {.callout-tip}
Groups with no matching NHN sheets get empty `flowlines.gpkg`/`waterbodies.gpkg` and a warning, then continue with `burn_streams = FALSE` — a missing NHN download never hard-stops the workflow.
:::

## Stage 4 — WhiteboxTools conditioning

Five steps run per group: breach depressions (least-cost, `dist = 10` cells, `fill = TRUE` to close remaining pits), D8 pointer, D8 flow accumulation (**must** take the pointer, not the DEM, with `pntr = TRUE` — passing the DEM silently re-derives flow direction and bypasses the stream-burned conditioning), stream extraction by threshold, and hillshade.

```{r}
#| label: nbe1-flow-accum
#| fig-cap: "D8 flow pointer (left) and flow accumulation (right) over the NBE1 unclipped catchment. These are Stage 5's per-site outputs (flow_accum.tif/flow_pointer.tif, not the Stage 7 *_clipped versions) — Stage 6/7's clipped versions cover only the much smaller downstream wedge that survives upstream removal (see Stage 6), which would leave most of this extent NA."
fa   <- rast("output/CELESTE/NBE1/flow_accum.tif")
ptr  <- rast("output/CELESTE/NBE1/flow_pointer.tif")
strm <- st_read("output/CELESTE/NBE1/streams.gpkg", quiet = TRUE)

dir_levels <- data.frame(ID = c(1, 2, 4, 8, 16, 32, 64, 128),
                          Direction = c("E", "SE", "S", "SW", "W", "NW", "N", "NE"))
levels(ptr) <- dir_levels

p_ptr <- ggplot() +
  tidyterra::geom_spatraster(data = ptr) +
  scale_fill_manual(values = pal_dir, name = "Flow to", na.translate = FALSE) +
  geom_sf(data = nbe1_catchment, fill = NA, colour = "black", linewidth = 0.5) +
  coord_sf(expand = FALSE) + theme_cw() +
  labs(title = "D8 flow pointer")

p_fa <- ggplot() +
  tidyterra::geom_spatraster(data = log10(fa + 1)) +
  scale_fill_viridis_c(option = "mako", name = "log10(cells)", na.value = NA) +
  geom_sf(data = strm, colour = "white", linewidth = 0.3) +
  geom_sf(data = nbe1_catchment, fill = NA, colour = "#E69F00", linewidth = 0.6) +
  coord_sf(expand = FALSE) + theme_cw() +
  labs(title = "Flow accumulation", subtitle = "White = extracted streams")

p_ptr + p_fa
```

```{r}
#| label: whitebox-code
#| eval: false
#| code-fold: false
whitebox::wbt_breach_depressions_least_cost(dem = dem_burned, output = dem_breached, dist = 10, fill = TRUE)
whitebox::wbt_d8_pointer(dem = dem_breached, output = flow_pointer)
whitebox::wbt_d8_flow_accumulation(input = flow_pointer, output = flow_accum, out_type = "cells", pntr = TRUE)  # <- pntr = TRUE is critical
whitebox::wbt_extract_streams(flow_accum = flow_accum, output = streams, threshold = stream_threshold)
whitebox::wbt_hillshade(dem = dem_breached, output = hillshade)
```

## Stage 5 — Delineate sites

Per site: write the pour point, snap it onto the nearest stream cell (`wbt_jenson_snap_pour_points()`, within `snap_dist`), delineate the watershed from the snapped point and D8 pointer, convert to polygon, then clip every group raster and the NHN flowlines to the result.

```{r}
#| label: nbe1-delineation
#| fig-cap: "NBE1: raw vs. snapped pour point over the full catchment (left), and a close-up of the snap onto the extracted stream network (right)."
pp_raw     <- st_read("output/CELESTE/NBE1/pour_point.gpkg", quiet = TRUE)
pp_snapped <- st_read("output/CELESTE/NBE1/pour_point_snapped.shp", quiet = TRUE)
dem1       <- rast("output/CELESTE/NBE1/dem.tif")
hill1      <- rast("output/CELESTE/NBE1/hillshade.tif")

p_full <- plot_relief(hill1, dem1, title = "Unclipped catchment") +
  geom_sf(data = nbe1_catchment, fill = NA, colour = "white", linewidth = 0.7) +
  geom_sf(data = pp_raw, aes(shape = "raw"), colour = "#D55E00", size = 2.6) +
  geom_sf(data = pp_snapped, aes(shape = "snapped"), colour = "#0072B2", size = 2.6) +
  scale_shape_manual(values = c(raw = 4, snapped = 16), name = NULL)

# Zoomed-in detail of the snap itself, against the (group-extent) streams
# raster it snapped onto — cropped to a small window so this stays fast.
snap_window <- st_buffer(pp_snapped, 250)
streams_tmp <- rast("output/CELESTE/NBE1/streams_tmp.tif") |> terra::crop(vect(snap_window))
snap_bb     <- st_bbox(snap_window)

# White-fill/black-outline marker for the snapped point — a solid blue dot
# (matching pp_snapped's colour everywhere else) all but disappears against
# the blue stream fill at this zoom, so contrast wins over colour-consistency
# for this one panel.
p_zoom <- ggplot() +
  tidyterra::geom_spatraster(data = streams_tmp, show.legend = FALSE) +
  scale_fill_gradient(low = "#1f78b4", high = "#1f78b4", na.value = "grey92") +
  geom_segment(aes(x = st_coordinates(pp_raw)[, 1], y = st_coordinates(pp_raw)[, 2],
                    xend = st_coordinates(pp_snapped)[, 1], yend = st_coordinates(pp_snapped)[, 2]),
               colour = "grey20", linetype = "dashed", linewidth = 0.4,
               arrow = arrow(length = unit(0.15, "cm"))) +
  geom_sf(data = pp_raw, colour = "#D55E00", shape = 4, size = 3, stroke = 1.3) +
  geom_sf(data = pp_snapped, shape = 21, fill = "white", colour = "black", size = 3, stroke = 1.1) +
  coord_sf(xlim = c(snap_bb["xmin"], snap_bb["xmax"]), ylim = c(snap_bb["ymin"], snap_bb["ymax"]), expand = FALSE) +
  theme_cw() +
  theme(axis.text = element_blank(), axis.ticks = element_blank()) +
  labs(title = "Snap detail", x = NULL, y = NULL,
       subtitle = "Blue = stream cells\nOrange x = raw pour point, white/black dot = snapped")

p_full + p_zoom + plot_layout(widths = c(1.3, 1))
```

`wbt_jenson_snap_pour_points()` moves the raw point onto the nearest cell of the **binary streams raster** (not the raw flow accumulation) within `snap_dist` metres — snapping against flow accumulation directly risks jumping to a distant, unrelated high-accumulation cell instead of the intended nearby channel.

```{r}
#| label: correcting-pour-point
#| eval: false
#| code-fold: false
# If a catchment is obviously wrong, the pour point likely snapped to the
# wrong stream cell. Open pour_point_snapped.shp + streams_tmp.tif in QGIS,
# drag the point onto the correct stream cell, then re-run watershed delineation
# only (skips re-snapping, reads the edited shapefile directly). This example
# predates the engine migration; the pre-engine mechanism shown at the time
# (workflow/R/stream/99_rerun_sites's rerun_site_watershed()) has since been
# retired along with the verification script that was its only remaining
# caller (see git history) — every active pipeline, including the current
# engine-based run_celeste.R, now uses workflow/R/engine/99_rerun_sites's
# rerun_engine_site_watershed() instead (same idea, CRS-dynamic):
source(here("workflow/R/engine/99_rerun_sites"))
rerun_engine_site_watershed("NBE1", sites, group_manifest, config, output_dir)
```

## Stage 6 — Remove upstream (nested catchments)

For sites where one pour point sits downstream of another (either within the same group or a shared network), the smaller upstream catchment(s) are erased from the larger one so each site's catchment reports contributing area only, without double-counting area that already belongs to an upstream site. `NBE1` is a strong real example: 10 upstream catchments were erased, taking its area from 236.9 km² down to 17.7 km².

```{r}
#| label: nbe-upstream-removal
#| fig-cap: "Unclipped vs. clipped area for all 8 NBE sites — the gap is area erased because it belongs to an upstream site."
nbe_nesting <- st_read("output/CELESTE/all_catchments_clipped.gpkg", quiet = TRUE) |>
  st_drop_geometry() |>
  dplyr::filter(str_detect(site_id, "^NBE[0-9]$")) |>
  select(site_id, area_km2_before, area_km2_after, n_erased) |>
  pivot_longer(c(area_km2_before, area_km2_after), names_to = "stage", values_to = "area_km2") |>
  mutate(stage = if_else(stage == "area_km2_before", "unclipped", "clipped"))

ggplot(nbe_nesting, aes(x = fct_rev(site_id), y = area_km2, fill = stage)) +
  geom_col(position = "identity", alpha = 0.85) +
  coord_flip() +
  scale_fill_manual(values = c(unclipped = "grey75", clipped = "#0072B2"), name = NULL) +
  theme_cw() +
  labs(title = "NBE: unclipped (grey, full bar) vs. clipped (blue) catchment area",
       x = NULL, y = expression("Area ("*km^2*")"))
```

```{r}
#| label: nbe-nesting-table
#| tbl-cap: "Upstream catchments erased per NBE site."
nbe_nesting_tbl <- st_read("output/CELESTE/all_catchments_clipped.gpkg", quiet = TRUE) |>
  st_drop_geometry() |>
  dplyr::filter(str_detect(site_id, "^NBE[0-9]$")) |>
  arrange(site_id)

gt(nbe_nesting_tbl) |>
  cols_label(site_id = "Site", area_km2_before = "Unclipped (km²)",
             area_km2_after = "Clipped (km²)", n_erased = "Catchments erased") |>
  fmt_number(c(area_km2_before, area_km2_after), decimals = 1) |>
  tab_header(title = "NBE upstream-removal summary")
```

```{r}
#| label: remove-upstream-code
#| eval: false
#| code-fold: false
group_manifest <- remove_upstream_catchments(group_manifest, sites, output_dir)
# -> writes catchment_clipped.gpkg per site, plus all_catchments_clipped.gpkg
```

## Stage 7 — Reclip outputs

Every raster and the flowlines are re-clipped to `catchment_clipped.gpkg`, producing the `*_clipped.tif` / `streams_clipped.gpkg` files used from here on.

```{r}
#| label: nbe1-clipped
#| fig-cap: "NBE1 after clipping — compare against the unclipped panel in Stage 5."
nbe1_clip     <- st_read("output/CELESTE/NBE1/catchment_clipped.gpkg", quiet = TRUE)
dem1_clip     <- rast("output/CELESTE/NBE1/dem_clipped.tif")
hill1_clip    <- terra::shade(terra::terrain(dem1_clip, "slope", unit = "radians"),
                               terra::terrain(dem1_clip, "aspect", unit = "radians"))

plot_relief(hill1_clip, dem1_clip, title = "NBE1 clipped catchment (17.7 km²)") +
  geom_sf(data = nbe1_clip, fill = NA, colour = "white", linewidth = 0.9) +
  geom_sf(data = pp_snapped, colour = "#0072B2", size = 2.8)
```

## Stage 8 — Morphometric metrics

Geometry, areal (shape), and relief parameters follow Shekar & Mathew (2024), computed for both catchment versions.

```{r}
#| label: nbe-metrics-table
#| tbl-cap: "Selected clipped-catchment metrics for the NBE group."
celeste_metrics <- read_csv("output/CELESTE/catchment_metrics.csv", show_col_types = FALSE)

nbe_metrics <- celeste_metrics |>
  dplyr::filter(str_detect(site_id, "^NBE[0-9]$"), version == "clipped") |>
  select(site_id, area_km2, form_factor, circularity_ratio, elongation_ratio,
         basin_relief_m, mean_slope_deg, drainage_density_km_km2, hypsometric_integral) |>
  arrange(site_id)

gt(nbe_metrics) |>
  cols_label(site_id = "Site", area_km2 = "Area (km²)", form_factor = "Form factor",
             circularity_ratio = "Circularity", elongation_ratio = "Elongation",
             basin_relief_m = "Relief (m)", mean_slope_deg = "Mean slope (°)",
             drainage_density_km_km2 = "Drainage density (km/km²)",
             hypsometric_integral = "Hypsometric integral") |>
  fmt_number(c(area_km2, form_factor, circularity_ratio, elongation_ratio, mean_slope_deg,
               drainage_density_km_km2, hypsometric_integral), decimals = 2) |>
  fmt_number(basin_relief_m, decimals = 0) |>
  tab_header(title = "NBE clipped-catchment morphometrics")
```

```{r}
#| label: nbe-area-vs-relief
#| fig-cap: "Basin relief vs. area for the NBE group (clipped catchments)."
ggplot(nbe_metrics, aes(area_km2, basin_relief_m, colour = site_id, label = site_id)) +
  geom_point(size = 3) +
  ggrepel::geom_text_repel(show.legend = FALSE, size = 3.2) +
  scale_colour_manual(values = pal_nbe, name = "Site") +
  scale_x_log10() +
  theme_cw() +
  labs(title = "NBE catchments: relief vs. area", x = expression("Area, log scale ("*km^2*")"), y = "Basin relief (m)")
```

Two of the metrics above are easiest to understand as pictures rather than formulas — both illustrated below directly from NBE1's real geometry and DEM.

```{r}
#| label: nbe1-shape-diagram
#| fig-cap: "NBE1 (clipped catchment), illustrating two areal metrics: the red diagonal is basin length Lb (diagonal of the axis-aligned bounding box, used by form_factor = A/Lb^2 and elongation_ratio); the dashed blue circle has the same area as the catchment, the reference used by circularity_ratio = A / area of that circle."
bb <- st_bbox(nbe1_clip)
diag_line <- st_sfc(
  st_linestring(matrix(c(bb["xmin"], bb["ymin"], bb["xmax"], bb["ymax"]), ncol = 2, byrow = TRUE)),
  crs = st_crs(nbe1_clip)
)
bbox_poly <- st_as_sfc(bb)

# st_point_on_surface(), not st_centroid() — a centroid can land outside a
# concave polygon (see CLAUDE.md's terra/hydroweight quirks); doesn't matter
# for this illustration, but it's the safe habit either way.
ctr     <- st_point_on_surface(nbe1_clip)
area_m2 <- as.numeric(st_area(nbe1_clip))
circle  <- st_buffer(ctr, sqrt(area_m2 / pi))

p_shape <- ggplot() +
  geom_sf(data = nbe1_clip, fill = "grey90", colour = "grey30") +
  geom_sf(data = bbox_poly, fill = NA, colour = "grey60", linetype = "dotted") +
  geom_sf(data = diag_line, colour = "#D55E00", linewidth = 0.9) +
  geom_sf(data = circle, fill = NA, colour = "#0072B2", linewidth = 0.9, linetype = "dashed") +
  theme_cw() +
  labs(title = "Shape metrics")

# Hypsometric curve — relative area (a/A) vs. relative height (h/H), computed
# directly from the clipped DEM's own elevation distribution. Its area under
# the curve approximates the table's hypsometric_integral for NBE1.
elev <- terra::values(dem1_clip, na.rm = TRUE) |> as.numeric() |> sort(decreasing = TRUE)
hyps <- tibble(
  rel_area   = seq_along(elev) / length(elev),
  rel_height = (elev - min(elev)) / (max(elev) - min(elev))
)

p_hyps <- ggplot(hyps, aes(rel_area, rel_height)) +
  geom_ribbon(aes(ymin = 0, ymax = rel_height), fill = "#0072B2", alpha = 0.15) +
  geom_line(colour = "#0072B2", linewidth = 0.9) +
  theme_cw() +
  labs(title = "Hypsometric curve", x = "Relative area (a/A)", y = "Relative height (h/H)")

p_shape + p_hyps
```

Reference: Shekar, P.R. & Mathew, A. (2024). Morphometric analysis of watersheds: a comprehensive review of data sources, quality, and geospatial techniques. *Watershed Ecology and the Environment*, 6, 13–25.

## Stage 9 — Hydroweight distance-weighted attributes

`target_O` for the stream pipeline is the snapped **pour point** (`output/<site_id>/pour_point.gpkg`) — a point has no `remove_region`, unlike the lake case below. DEM/flow accumulation are resolved per site from that site's own group cache, since stream groups (NIP/TUR/KEN in Ontario, NBE/COC in New Brunswick, MOR in PEI) span independent HydroBasins regions. Each of seven weighting schemes distance-weights every LOI value by its position relative to the pour point:

```{r}
#| label: weighting-schemes-table
#| tbl-cap: "hydroweight weighting schemes."
schemes <- tribble(
  ~Scheme, ~Description,
  "lumped",  "Unweighted areal mean/proportion — every cell counted equally",
  "iEucO",   "Inverse Euclidean distance to the outflow (O) target",
  "iFLO",    "Inverse flow-path distance to the outflow, along the D8 network",
  "HAiFLO",  "iFLO further weighted by hydrologically-active area (flow accumulation)",
  "iEucS",   "Inverse Euclidean distance to the nearest stream cell",
  "iFLS",    "Inverse flow-path distance to the nearest stream cell",
  "HAiFLS",  "iFLS further weighted by hydrologically-active area"
)
gt(schemes)
```

Three LOIs are prepared for the stream pipeline (`workflow/CELESTE/prepare_*.R`): continuous NDVI (annual mosaic, 1984–2025), a per-pixel NDVI Theil-Sen slope/Mann-Kendall trend (one raster per group, ~29 min one-time), and a categorical harvest/regen layer (Ontario + New Brunswick sources, 2002–2024, one band per year plus a combined band).

```{r}
#| label: nbe-hydroweight
#| fig-cap: "NBE: long-term NDVI trend (Theil-Sen slope, unweighted zonal mean) vs. proportion of catchment ever harvested 2002-2024 (areal/lumped), clipped catchments."
celeste_hw <- read_csv("output/CELESTE/CELESTE_hydroweight.csv", show_col_types = FALSE)

nbe_hw <- celeste_hw |>
  dplyr::filter(str_detect(site, "^NBE[0-9]$"), version == "clipped") |>
  select(site, ndvi_trend_slope_mean, harvest_regen_combined_harvest_lumped_prop) |>
  arrange(site)

p1 <- ggplot(nbe_hw, aes(fct_rev(site), ndvi_trend_slope_mean, fill = site)) +
  geom_col(show.legend = FALSE) +
  geom_hline(yintercept = 0, linewidth = 0.3) +
  coord_flip() +
  scale_fill_manual(values = pal_nbe) +
  theme_cw() +
  labs(title = "NDVI trend (Sen's slope/yr)", x = NULL, y = "Slope")

p2 <- ggplot(nbe_hw, aes(fct_rev(site), harvest_regen_combined_harvest_lumped_prop, fill = site)) +
  geom_col(show.legend = FALSE) +
  coord_flip() +
  scale_y_continuous(labels = percent_format()) +
  scale_fill_manual(values = pal_nbe) +
  theme_cw() +
  labs(title = "Ever harvested 2002-2024", x = NULL, y = "% of catchment")

p1 + p2
```

The 7 schemes above aren't interchangeable — they can give visibly different answers for the same site and the same LOI, depending on where within the catchment the disturbed area actually sits relative to the outflow:

```{r}
#| label: nbe-schemes-compare
#| fig-cap: "How the choice of weighting scheme changes the answer: proportion of catchment classed as harvested (2002-2024), per NBE site, under all 7 schemes."
scheme_labels <- c(lumped = "lumped", ieuco = "iEucO", iflo = "iFLO", haiflo = "HAiFLO",
                    ieucs = "iEucS", ifls = "iFLS", haifls = "HAiFLS")

nbe_schemes <- celeste_hw |>
  dplyr::filter(str_detect(site, "^NBE[0-9]$"), version == "clipped") |>
  select(site, matches(glue::glue("^harvest_regen_combined_harvest_({paste(names(scheme_labels), collapse = '|')})_prop$"))) |>
  pivot_longer(-site, names_to = "scheme", values_to = "prop") |>
  mutate(
    scheme = str_remove(scheme, "^harvest_regen_combined_harvest_") |> str_remove("_prop$"),
    scheme = factor(scheme_labels[scheme], levels = scheme_labels)
  )

ggplot(nbe_schemes, aes(scheme, prop, group = site, colour = site)) +
  geom_line(linewidth = 0.6, alpha = 0.8) +
  geom_point(size = 1.8) +
  scale_colour_manual(values = pal_nbe, name = "Site") +
  scale_y_continuous(labels = percent_format()) +
  theme_cw() +
  theme(axis.text.x = element_text(angle = 30, hjust = 1)) +
  labs(title = "Harvested proportion under each weighting scheme", x = NULL, y = "% of catchment")
```

```{r}
#| label: nbe-interactive-map
#| cache: false
#| fig-cap: "Interactive: NBE clipped catchments and pour points (HTML only — a static overview replaces this in the PDF)."
nbe_catchments_clip <- st_read("output/CELESTE/all_catchments_clipped.gpkg", quiet = TRUE) |>
  dplyr::filter(str_detect(site_id, "^NBE[0-9]$"))

if (is_html) {
  suppressPackageStartupMessages(library(tmap))
  tmap_mode("view")
  tm_shape(nbe_catchments_clip) +
    tm_polygons(fill = "site_id", fill.scale = tm_scale(values = pal_nbe), fill_alpha = 0.5,
                fill.legend = tm_legend(title = "Site")) +
    tm_shape(nbe_pour_points) +
    tm_dots(size = 0.6, fill = "black")
} else {
  ggplot() +
    geom_sf(data = nbe_catchments_clip, aes(fill = site_id), colour = "white", alpha = 0.75) +
    geom_sf(data = nbe_pour_points, size = 1.6) +
    scale_fill_manual(values = pal_nbe, name = "Site") +
    theme_cw() +
    labs(title = "NBE clipped catchments overview")
}
```

```{r}
#| label: hydroweight-code-stream
#| eval: false
#| code-fold: false
hw_results <- calculate_hydroweight_attributes_stream(
  sites          = sites,
  group_manifest = group_manifest,
  output_dir     = output_dir,
  cache_dir      = cache_dir,
  loi_layers     = list(
    ndvi          = list(name = "ndvi", type = "continuous", path_lazy = prepare_ndvi_vrt()),
    ndvi_trend    = list(name = "ndvi_trend", type = "continuous", path_template = "…/{group_id}_ndvi_trend.tif"),
    harvest_regen = list(name = "harvest_regen", type = "categorical", path_lazy = "…", class_levels = harvest_regen_levels)
  )
)
```

---

# Worked example B — Lake delineation (`Clearwater`, `Swan`, `Tilton`, `Lohi`)

These 4 lakes share one project-wide OIH DEM/flow-accumulation raster (`cache/CAM/`) rather than a per-group one — the lake pipeline doesn't span disjoint HydroBasins regions the way the stream pipeline's groups do, so there's no equivalent of the group manifest here.

## Stage 1 — Match lake polygons

`match_lake_polygons()` matches each lake site to an OHN waterbody polygon via a three-pass strategy: spatial buffer intersection, fuzzy name matching (Jaro-Winkler against `OFFICIAL_N`), then manual `OGF_ID` lookup for confirmed hard cases. The matched polygon is rasterized (buffered 30 m) to the full D8 extent as the pour-point target for `wbt_watershed()` — the map below reconstructs each lake's footprint from that rasterized pour-point target, over an OpenStreetMap basemap for geographic context.

```{r}
#| label: cam-lakes-map
#| cache: false
#| fig-cap: "Interactive: the 4 example lakes on an OpenStreetMap basemap (HTML only — small-multiple outlines replace this in the PDF). Basemap tiles are fetched live by the viewer's browser, not embedded at render time."
cam_lakes <- c("Clearwater", "Swan", "Tilton", "Lohi")

lake_outline <- function(lake) {
  r <- rast(glue::glue("output/CAM/lake_delineation/{lake}/lake_pourpoint.tif")) |> terra::trim()
  terra::as.polygons(r, dissolve = TRUE) |> st_as_sf() |> mutate(site_id = lake)
}
cam_lake_polys <- map(cam_lakes, lake_outline) |> bind_rows()

if (is_html) {
  suppressPackageStartupMessages(library(tmap))
  tmap_mode("view")
  tm_basemap("OpenStreetMap") +
    tm_shape(cam_lake_polys) +
    tm_polygons(fill = "site_id", fill.scale = tm_scale(values = pal_lakes),
                fill.legend = tm_legend(title = "Lake"))
} else {
  # facet_wrap(scales = "free") isn't supported with coord_sf(), and these 4
  # lakes are geographically far apart — so each gets its own small map (own
  # extent) combined with patchwork, rather than one shared-extent facet grid.
  lake_maps <- map(cam_lakes, \(lake) {
    ggplot(dplyr::filter(cam_lake_polys, site_id == lake)) +
      geom_sf(fill = pal_lakes[[lake]], colour = "grey20") +
      theme_cw() +
      theme(axis.text = element_blank(), legend.position = "none") +
      labs(title = lake)
  })
  wrap_plots(lake_maps, nrow = 1) +
    plot_annotation(title = "Matched lake polygons (from rasterized pour-point target)")
}
```

```{r}
#| label: match-lakes-code
#| eval: false
#| code-fold: false
lake_polys <- match_lake_polygons(
  sites            = sites,
  lakes_path       = OIH_LAKES_PATH,
  buffer_m         = 10,
  name_dist_max    = 0.15,
  name_search_km   = 40
)
```

## Stage 2 — Prepare OIH DEM and flow products

One project-wide DEM and D8 flow-accumulation raster are prepared for the whole of the OIH-covered Ontario domain (not per group, unlike the stream pipeline) — `cache/CAM/dem.tif`, `d8_pntr.tif`, `flow_accum.tif`.

```{r}
#| label: clearwater-relief
#| fig-cap: "Clearwater lake catchment: relief, from its own clipped rasters. The lake is shown as an outline (not filled) so the DEM underneath it stays visible."
cw_dem   <- rast("output/CAM/lake_delineation/Clearwater/dem.tif")
cw_hill  <- terra::shade(terra::terrain(cw_dem, "slope", unit = "radians"),
                          terra::terrain(cw_dem, "aspect", unit = "radians"))
cw_catch <- st_read("output/CAM/lake_delineation/Clearwater/catchment.gpkg", quiet = TRUE)

plot_relief(cw_hill, cw_dem, title = "Clearwater catchment") +
  geom_sf(data = cw_catch, fill = NA, colour = "white", linewidth = 0.8) +
  geom_sf(data = st_as_sf(cam_lake_polys[cam_lake_polys$site_id == "Clearwater", ]),
          fill = NA, colour = "#0072B2", linewidth = 1)
```

## Stage 3 — Delineate lake catchments

The matched (buffered) lake polygon is rasterized as the pour-point target and passed to `wbt_watershed()` — analogous to the snapped point in the stream pipeline, but an area rather than a single cell.

```{r}
#| label: cam-4lakes-catchments
#| fig-cap: "Unclipped catchments for the 4 example lakes. Each is outlined in black so overlapping/adjacent boundaries stay legible even where fills are similar."
cam_catchments <- st_read("output/CAM/lake_delineation/all_catchments.gpkg", quiet = TRUE) |>
  dplyr::filter(site_id %in% cam_lakes)

ggplot(cam_catchments) +
  geom_sf(aes(fill = site_id), colour = "black", linewidth = 0.4, alpha = 0.85) +
  geom_sf(data = cam_lake_polys, fill = "grey15", colour = NA) +
  scale_fill_manual(values = pal_lakes, name = "Lake") +
  theme_cw() +
  labs(title = "Clearwater, Swan, Tilton, Lohi: unclipped catchments", subtitle = "Dark fill = lake footprint")
```

```{r}
#| label: delineate-lakes-code
#| eval: false
#| code-fold: false
results <- delineate_lake_catchments(
  sites      = sites,
  lake_polys = lake_polys,
  cache_dir  = cache_dir,
  output_dir = output_dir
)
```

## Stage 4 — Remove upstream lakes

Same nesting concept as Stage 6 for streams, applied to lake catchments. Among these 4 lakes, `Swan`'s clipped area is identical to its unclipped area (no upstream lake to erase), while `Lohi` loses most of its area to upstream removal — the interactive map below lets you toggle the unclipped and clipped layers to see exactly which lakes sit upstream of which.

```{r}
#| label: cam-4lakes-nesting
#| fig-cap: "Unclipped vs. clipped area for the 4 example lakes (from catchment_metrics.csv)."
cam_metrics <- read_csv("output/CAM/lake_delineation/catchment_metrics.csv", show_col_types = FALSE)

cam_nesting <- cam_metrics |>
  dplyr::filter(site_id %in% cam_lakes) |>
  select(site_id, version, area_km2) |>
  pivot_wider(names_from = version, values_from = area_km2)

gt(cam_nesting) |>
  cols_label(site_id = "Lake", unclipped = "Unclipped (km²)", clipped = "Clipped (km²)") |>
  fmt_number(c(unclipped, clipped), decimals = 2) |>
  tab_header(title = "CAM upstream-removal effect, 4 example lakes")
```

```{r}
#| label: cam-nesting-interactive
#| cache: false
#| fig-cap: "Interactive: unclipped catchments (pale, full extent) vs. clipped catchments (solid, black outline) for the 4 example lakes — use the layer control to toggle each on/off and see how the smaller lakes nest inside the larger ones. HTML only."
cam_catchments_unclip <- st_read("output/CAM/lake_delineation/all_catchments.gpkg", quiet = TRUE) |>
  dplyr::filter(site_id %in% cam_lakes)
cam_catchments_clip4 <- st_read("output/CAM/lake_delineation/all_catchments_clipped.gpkg", quiet = TRUE) |>
  dplyr::filter(site_id %in% cam_lakes)

if (is_html) {
  suppressPackageStartupMessages(library(tmap))
  tmap_mode("view")
  tm_shape(cam_catchments_unclip) +
    tm_polygons(fill = "site_id", fill.scale = tm_scale(values = pal_lakes), fill_alpha = 0.25,
                fill.legend = tm_legend(show = FALSE), group = "Unclipped catchments") +
    tm_shape(cam_catchments_clip4) +
    tm_polygons(fill = "site_id", fill.scale = tm_scale(values = pal_lakes), fill_alpha = 0.7,
                col = "black", lwd = 1, fill.legend = tm_legend(title = "Lake"), group = "Clipped catchments") +
    tm_shape(cam_lake_polys) +
    tm_polygons(fill = "grey15", fill.legend = tm_legend(show = FALSE), group = "Lakes")
} else {
  ggplot() +
    geom_sf(data = cam_catchments_unclip, aes(fill = site_id), alpha = 0.25, colour = NA) +
    geom_sf(data = cam_catchments_clip4, aes(fill = site_id), colour = "black", alpha = 0.7) +
    geom_sf(data = cam_lake_polys, fill = "grey15", colour = NA) +
    scale_fill_manual(values = pal_lakes, name = "Lake") +
    theme_cw() +
    labs(title = "Unclipped (pale) vs. clipped (solid) catchments")
}
```

```{r}
#| label: remove-upstream-lake-code
#| eval: false
#| code-fold: false
lake_polys <- remove_upstream_lake_catchments(lake_polys, sites, cache_dir, output_dir)
```

## Stage 5 — Reclip outputs

Same `reclip_outputs.R` module as the stream pipeline — it works from either a `group_manifest` (stream) or a `cache_dir` (lake), producing the same `*_clipped.tif` / `streams_clipped.gpkg` convention either way.

## Stage 6 — Morphometric metrics

The same `catchment_metrics.R` module also reports lake-specific morphometric parameters (fetch, shoreline development, lake area %, catchment-area:lake-area ratio) alongside the standard geometry/areal/relief set, when a lake polygon is present.

```{r}
#| label: cam-metrics-table
#| tbl-cap: "Selected clipped-catchment metrics, 4 example lakes."
cam_metrics_tbl <- cam_metrics |>
  dplyr::filter(site_id %in% cam_lakes, version == "clipped") |>
  select(site_id, area_km2, lake_area_ha, ca_la_ratio, shoreline_development,
         mean_slope_deg, drainage_density_km_km2) |>
  arrange(site_id)

gt(cam_metrics_tbl) |>
  cols_label(site_id = "Lake", area_km2 = "Catchment area (km²)", lake_area_ha = "Lake area (ha)",
             ca_la_ratio = "CA:LA ratio", shoreline_development = "Shoreline dev.",
             mean_slope_deg = "Mean slope (°)", drainage_density_km_km2 = "Drainage density") |>
  fmt_number(everything(), decimals = 2) |>
  tab_header(title = "CAM clipped-catchment morphometrics")
```

These 4 lakes were picked as examples because they genuinely span a wide range on most of these metrics — from `Swan`'s tiny, near-1:1 catchment to `Lohi`'s much larger, higher-ratio one. As in Stage 8 for the stream pipeline, that variation is easier to read as pictures than as table rows: a shape diagram per lake, and their hypsometric curves overlaid.

```{r}
#| label: cam-shape-diagrams
#| fig-cap: "Shape metrics for each of the 4 example lake catchments: the coloured diagonal is basin length Lb; the dashed circle is the equal-area circle used by circularity_ratio. Each lake gets its own extent (small multiples, as in Stage 1) since they're geographically far apart."
cam_shape_diagram <- function(lake) {
  catch <- st_read(glue::glue("output/CAM/lake_delineation/{lake}/catchment_clipped.gpkg"), quiet = TRUE)

  bb <- st_bbox(catch)
  diag_line <- st_sfc(
    st_linestring(matrix(c(bb["xmin"], bb["ymin"], bb["xmax"], bb["ymax"]), ncol = 2, byrow = TRUE)),
    crs = st_crs(catch)
  )
  bbox_poly <- st_as_sfc(bb)

  ctr     <- st_point_on_surface(catch)
  area_m2 <- as.numeric(st_area(catch))
  circle  <- st_buffer(ctr, sqrt(area_m2 / pi))

  ggplot() +
    geom_sf(data = catch, fill = "grey90", colour = "grey30") +
    geom_sf(data = bbox_poly, fill = NA, colour = "grey60", linetype = "dotted") +
    geom_sf(data = diag_line, colour = pal_lakes[[lake]], linewidth = 1) +
    geom_sf(data = circle, fill = NA, colour = "black", linewidth = 0.7, linetype = "dashed") +
    theme_cw() +
    theme(axis.text = element_blank()) +
    labs(title = lake)
}

wrap_plots(map(cam_lakes, cam_shape_diagram), nrow = 1)
```

```{r}
#| label: cam-hypsometric
#| fig-cap: "Hypsometric curves for the 4 example lake catchments, computed directly from each lake's dem_clipped.tif."
cam_hypsometric <- function(lake) {
  dem  <- rast(glue::glue("output/CAM/lake_delineation/{lake}/dem_clipped.tif"))
  elev <- terra::values(dem, na.rm = TRUE) |> as.numeric() |> sort(decreasing = TRUE)
  tibble(
    site_id    = lake,
    rel_area   = seq_along(elev) / length(elev),
    rel_height = (elev - min(elev)) / (max(elev) - min(elev))
  )
}

cam_hyps <- map(cam_lakes, cam_hypsometric) |> bind_rows()

ggplot(cam_hyps, aes(rel_area, rel_height, colour = site_id)) +
  geom_line(linewidth = 0.9) +
  scale_colour_manual(values = pal_lakes, name = "Lake") +
  theme_cw() +
  labs(title = "Hypsometric curves", x = "Relative area (a/A)", y = "Relative height (h/H)")
```

## Stage 7 — Hydroweight distance-weighted attributes

`target_O` here is the **lake polygon** itself (not a point), so `remove_region` excises the lake's own footprint from LOI summaries — a stream pour point has no area to exclude. CAM's demonstrated LOI is CANLCC 2020 land cover, reprojected once to EPSG:3161 and cached (`cache/CAM/hydroweight_loi/canlcc_reprojected.tif`).

```{r}
#| label: cam-hydroweight
#| fig-cap: "Land cover composition of the 4 example lake catchments (lumped/areal weighting)."
cam_hw <- read_csv("output/CAM/lake_delineation/CAM_hydroweight.csv", show_col_types = FALSE)

cam_hw_long <- cam_hw |>
  dplyr::filter(site %in% cam_lakes) |>
  select(site, ends_with("_lumped_prop")) |>
  pivot_longer(-site, names_to = "class", values_to = "prop") |>
  mutate(class = str_remove(class, "^canlcc_") |> str_remove("_lumped_prop$") |> str_replace_all("_", " ") |> str_to_title())

ggplot(cam_hw_long, aes(x = site, y = prop, fill = class)) +
  geom_col(position = "stack") +
  scale_y_continuous(labels = percent_format()) +
  # 10 CANLCC classes — beyond the 8-colour Okabe-Ito set used elsewhere for
  # site/lake identity, so a larger qualitative palette is used here instead.
  scale_fill_brewer(palette = "Paired", name = "CANLCC class") +
  theme_cw() +
  labs(title = "CANLCC land cover composition (lumped)", x = NULL, y = "% of catchment")
```

```{r}
#| label: cam-interactive-map
#| cache: false
#| fig-cap: "Interactive: the 4 example lake catchments with the rasterized CANLCC land cover layer underneath (HTML only — a static overview replaces this in the PDF)."
cam_catchments_clip <- st_read("output/CAM/lake_delineation/all_catchments_clipped.gpkg", quiet = TRUE) |>
  dplyr::filter(site_id %in% cam_lakes)

if (is_html) {
  suppressPackageStartupMessages(library(tmap))

  # The full CANLCC raster is huge (61225 x 62182 cells); cropping first —
  # rather than reading it whole — is what keeps this fast (same pattern
  # raster_attributes.R uses for any path_lazy LOI source).
  canlcc_levels_lookup <- data.frame(
    ID    = c(1L, 2L, 5L, 6L, 8L, 10L, 11L, 12L, 13L, 14L, 15L, 16L, 17L, 18L, 19L),
    Class = c("needleleaf_forest", "taiga_needleleaf_forest", "broadleaf_deciduous_forest",
              "mixed_forest", "temperate_shrubland", "temperate_grassland", "shrubland_lichen_moss",
              "grassland_lichen_moss", "barren_lichen_moss", "wetland", "cropland", "barren",
              "urban", "water", "snow_ice")
  )
  canlcc_bb   <- st_bbox(st_buffer(cam_catchments_clip, 500))
  canlcc_crop <- rast("cache/CAM/hydroweight_loi/canlcc_reprojected.tif") |>
    terra::crop(terra::ext(canlcc_bb["xmin"], canlcc_bb["xmax"], canlcc_bb["ymin"], canlcc_bb["ymax"]))
  levels(canlcc_crop) <- canlcc_levels_lookup

  tmap_mode("view")
  tm_shape(canlcc_crop) +
    tm_raster(col.legend = tm_legend(title = "Land cover")) +
    tm_shape(cam_catchments_clip) +
    tm_polygons(fill = "site_id", fill.scale = tm_scale(values = pal_lakes), fill_alpha = 0.35,
                col = "black", lwd = 1.5, fill.legend = tm_legend(title = "Lake"))
} else {
  ggplot() +
    geom_sf(data = cam_catchments_clip, aes(fill = site_id), colour = "black", alpha = 0.75) +
    geom_sf(data = cam_lake_polys, fill = "grey15", colour = NA) +
    scale_fill_manual(values = pal_lakes, name = "Lake") +
    theme_cw() +
    labs(title = "CAM clipped catchments overview")
}
```

```{r}
#| label: hydroweight-code-lake
#| eval: false
#| code-fold: false
hw_results <- calculate_hydroweight_attributes(
  sites      = sites,
  lake_polys = lake_polys,
  cache_dir  = cache_dir,
  output_dir = output_dir,
  loi_layers = list(
    canlcc = list(name = "canlcc", type = "categorical", path = CANLCC_PATH, class_levels = canlcc_levels)
  )
)
```

---

# Cross-cutting concepts

## LOI layer descriptor shapes

`loi_layers[[i]]` supports exactly one of three source shapes, chosen per LOI based on how the source data is distributed:

| Shape | When to use | Example |
|---|---|---|
| `path` | One raster already covering the full site extent | CANLCC (lake pipeline) |
| `path_template` | One pre-clipped raster per site, `"{site_id}"` placeholder | — |
| `path_lazy` | Shared source too large to reproject/cache whole (mosaic VRT, or one raster per `"{group_id}"`); cropped to each site's catchment *before* being reprojected or written to disk | NDVI mosaic VRT, NDVI trend, harvest/regen (stream pipeline) |

## Categorical LOI quirks (found the hard way)

A handful of `terra`/`hydroweight` behaviours are easy to get wrong silently when building a multi-band categorical LOI like harvest/regen. All were confirmed by direct testing while building that layer (24 bands: 23 years + combined):

- `terra::levels(r) <- df` on a **multi-layer** SpatRaster only applies to layer 1 — layers 2+ silently keep default auto-generated labels. Build each band as an independent single-layer raster, set its levels, then combine via `terra::rast(list(...))`; never mutate a multi-layer factor object in place afterwards.
- An exactly-3-band Byte GeoTIFF gets GDAL's automatic RGB interpretation on write, silently renaming bands to "red"/"green"/"blue" on every subsequent read — use `gdal = "PHOTOMETRIC=MINISBLACK"`.
- A class covering 100% of a catchment produces an id-less, degenerate column name — detected and handled directly (its distance-weighted proportion is provably 1.0 under any scheme) rather than passed through `hydroweight_attributes()`.
- A class with zero occurrences (but not the *only* class) is omitted from that row's output entirely, not reported as 0 — becomes a misleading `NA` after binding multiple sites unless explicitly back-filled.
- `sf::st_read()`'s `wkt_filter` is pushed down to OGR *before* R-side geometry cleanup, so a source with curved geometries can silently drop or entirely miss features depending on the source — safer to read the full layer, clean geometry, then filter spatially in R.

## Caching and incremental reruns

Every stage checks `cache_exists(path)` (file exists **and** is non-empty) before writing, so a rerun after a crash or a newly-added site safely skips completed work — only the new site triggers Stage 5+ (delineation onward); the group's DEM conditioning from Stages 2–4 is reused as-is.

```{r}
#| label: cache-exists-code
#| eval: false
#| code-fold: false
cache_exists <- function(path) fs::file_exists(path) && fs::file_size(path) > 0
```

To force specific parts to rerun, `code/reset_workflow.R` provides:

```{r}
#| label: reset-code
#| eval: false
#| code-fold: false
reset_site("NBE1")            # one site's output/ folder only
reset_group("NBE", sites)     # one group's cache + all its sites' outputs
reset_sites(sites)            # every site's outputs, keep the group cache
reset_all()                   # everything — cache and outputs
```

## Correcting a misaligned pour point / lake target

| Pipeline | Symptom | Fix |
|---|---|---|
| Stream | Catchment too small / wrong tributary | Edit `output/<site_id>/pour_point_snapped.shp` in QGIS against `streams_tmp.tif`, then `rerun_site_watershed()` — reads the edited shapefile directly, skips re-snapping |
| Lake | Catchment misses part of the lake, or grabs a neighbouring one | Edit that site's `lake_pourpoint.tif` in QGIS, then `rerun_watershed_lake()` — re-runs watershed from the edited raster without regenerating it from the polygon match |

---

# Troubleshooting

**Interior pit cells warning from WhiteboxTools.** The DEM was masked to the AOI polygon before conditioning (introducing NoData that reads as depressions). `prepare_dem()` crops to the rectangular extent only by design — check nothing external re-masked the DEM before this stage.

**Flow accumulation max value implausibly small.** `wbt_d8_flow_accumulation()` was passed the DEM instead of the flow pointer, or `pntr = TRUE` was omitted — verify both if calling WhiteboxTools functions directly outside the module.

**NHN sheet not found for a group.** The corresponding GDB hasn't been downloaded. The workflow degrades gracefully (empty flowlines/waterbodies, `burn_streams` effectively off for that group) rather than erroring — download the sheet from GeoBase NHN and place it in `nhn_dir` if burning is actually needed there.

**A stream group's LOI mosaic reads as 0 everywhere instead of NA.** `build_mosaic_vrt()` always sets `-vrtnodata` for exactly this reason — without it, any area not covered by *any* input tile reads as `0`, indistinguishable from real data. This was confirmed directly: before the fix, the NBE group (no NDVI tile coverage at the time) silently produced "NDVI = 0" for every NBE catchment instead of failing the "all-NA after crop/mask" check that would have excluded it correctly.

**GEOS/invalid-geometry errors from HydroBasins or waterbody sources.** `st_make_valid()` is applied inside `build_group_aoi()` (stream) and before rasterization in `rasterize_competing_classes()`. If the error persists it's likely a different geometry op — check the traceback for exactly which call raised it.

---

```{r}
#| label: session-info
#| code-fold: true
sessionInfo()
```
