---
title: "Catchment Delineation Workflow"
subtitle: "A practical guide to the stream and lake delineation pipelines"
author: "Sam Woodman — Natural Resources Canada"
date: today
format:
  html:
    toc: true
    toc-depth: 3
    toc-location: left
    toc-title: "Contents"
    code-fold: true
    code-tools: true
    code-link: true
    theme: flatly
    highlight-style: github
    embed-resources: true
    smooth-scroll: true
    number-sections: true
    fig-width: 7.5
    fig-height: 5.5
  pdf:
    toc: true
    toc-depth: 3
    number-sections: true
    documentclass: article
    fig-pos: "H"
    fig-width: 6.5
    fig-height: 4.8
    geometry:
      - margin=2.5cm
execute:
  echo: true
  warning: false
  message: false
  cache: false
---

```{r}
#| label: setup
#| include: false
#| cache: false

library(here)
library(sf)
library(terra)
library(dplyr)
library(tidyr)
library(purrr)
library(stringr)
library(forcats)
library(readr)
library(tibble)
library(ggplot2)
library(tidyterra)
library(gt)
library(scales)
library(patchwork)

# This document is read-only with respect to the workflow: every figure and
# table below loads files that already exist under cache/ and output/ from a
# completed run of workflow/CELESTE/run_celeste.R and workflow/run_cam_lakes.R.
# No delineation, conditioning, or hydroweight step is re-run to build it, so
# it renders in well under a minute rather than hours. Pipeline code shown
# alongside each stage (grey, non-executed blocks) is reference — copy it into
# an R session if you actually need to run that step.
#
# This file lives in docs/, one level below the project root, but every path
# below (cache/..., output/...) is written root-relative — set knitr's
# working directory to the project root once, here, rather than prefixing
# every read with "../".
knitr::opts_knit$set(root.dir = here::here())

# The stream worked example reads output/CELESTE/ and cache/CELESTE/NBE/ —
# the current, engine-based run_celeste.R's own output. (Pre-2026-08-31 this
# lived under a separate output/CELESTE_engine/ + cache/CELESTE_engine/,
# physically apart from a still-on-disk pre-engine run of the same name; a
# 2026-08-31 consolidation retired the pre-engine copy for real and renamed
# CELESTE_engine -> CELESTE. See git history around that date if older
# material still references the old paths.)
#
# This guide's worked examples (NBE for streams, 4 CAM lakes) cover the two
# *pipeline kinds* the engine supports — point pour point and lake polygon.
# Every project added since (CAM streams, EMILY_TURKEY) is one of those two
# kinds on the same engine, not a third kind, so they aren't walked through
# step-by-step here; the Overview and Cross-cutting sections below describe
# what differs about them instead. See each project's own
# workflow/<PROJECT>/README.md for full detail.
#
# Chunk caching (knitr's execute: cache:) is deliberately OFF: every chunk
# here just rereads small-to-moderate local files under cache/ and output/
# (no re-running of the GIS pipeline itself, no network calls other than
# maptiles::get_tiles()/rnaturalearth::ne_states(), both of which keep their
# own small on-disk tile/data caches independent of knitr — rosm.cache/ is
# ~64K). A full cold render takes ~50s; knitr caching only shaved that to
# ~22s by pickling large SpatRaster/sf R objects to disk, at a cost of
# ~32MB of *_cache/ that regenerates on every render and was never actually
# needed for speed. Turn it back on (cache: true in the YAML above) only if
# this document grows chunks that are genuinely expensive to recompute.
knitr::opts_chunk$set(fig.align = "center")

# -- Shared plotting helpers --------------------------------------------------

# Fixed-order, colourblind-safe base palette (Okabe & Ito, 2008). Named
# palettes are derived from it below and used (never re-cycled to a
# different meaning) so a given site/lake/direction keeps the same colour
# wherever it appears in the document — including across figures where a
# factor's sort order differs, since named lookup is used, not position.
pal_okabe_ito <- c(
  "#E69F00", "#56B4E9", "#009E73", "#F0E442",
  "#0072B2", "#D55E00", "#CC79A7", "#000000"
)

pal_nbe <- setNames(pal_okabe_ito, paste0("NBE", 1:8))

# Hand-picked (not positional) so none of the 4 are confusable — Okabe-Ito's
# two blues in particular are avoided here.
pal_lakes <- c(
  Clearwater = "#E69F00",
  Swan       = "#009E73",
  Tilton     = "#D55E00",
  Lohi       = "#CC79A7"
)

# WhiteboxTools' D8 pointer uses the ESRI encoding: 1=E, 2=SE, 4=S, 8=SW,
# 16=W, 32=NW, 64=N, 128=NE.
pal_dir <- setNames(pal_okabe_ito, c("E", "SE", "S", "SW", "W", "NW", "N", "NE"))

theme_cw <- function(base_size = 11) {
  theme_minimal(base_size = base_size) +
    theme(
      legend.position   = "bottom",
      panel.grid.minor  = element_blank(),
      plot.title        = element_text(face = "bold", size = rel(1.05)),
      plot.subtitle      = element_text(colour = "grey35"),
      strip.text         = element_text(face = "bold")
    )
}

# A hillshade + elevation "relief" figure is used repeatedly (group AOI,
# per-site DEM, per-lake DEM). Centralising it keeps every occurrence
# consistent and keeps individual chunks short.
plot_relief <- function(hillshade, dem = NULL, title = NULL, subtitle = NULL) {
  p <- ggplot() +
    tidyterra::geom_spatraster(data = hillshade, show.legend = FALSE) +
    scale_fill_gradient(low = "grey5", high = "grey95", na.value = NA)
  if (!is.null(dem)) {
    p <- p +
      ggnewscale::new_scale_fill() +
      tidyterra::geom_spatraster(data = dem, alpha = 0.55) +
      scale_fill_viridis_c(option = "viridis", name = "Elevation (m)", na.value = NA)
  }
  p + coord_sf(expand = FALSE) + theme_cw() +
    labs(title = title, subtitle = subtitle, x = NULL, y = NULL)
}

is_html <- knitr::is_html_output()
```

::: {.callout-note}
This guide was drafted with assistance from Claude (Anthropic), reading directly from this repository's code and cached outputs.
:::

## Overview

This workflow delineates catchments from pour-point coordinates for sites across Canada and computes morphometric and distance-weighted landscape metrics for each one. It is **project-agnostic**: a sites table and a handful of shared external datasets go in, self-contained per-site output folders come out.

Every project on it reduces to one of two **pipeline kinds**, sharing most of their machinery but differing wherever the feature type forces a difference — what the "outflow" target is, and whether flow products are prepared per-group or once for the whole project. This guide uses a real, already-processed example of each:

| | **Stream pipeline** | **Lake pipeline** |
|---|---|---|
| Feature type | River/stream pour point | Lake polygon |
| Example used below | `NBE` group — 8 New Brunswick sites (`NBE1`–`NBE8`) | `Clearwater`, `Swan`, `Tilton`, `Lohi` |
| Runner | `workflow/CELESTE/run_celeste.R` | `workflow/run_cam_lakes.R` |
| CRS | EPSG:3979 (Canada Atlas Lambert) | EPSG:3161 (Ontario MNR Lambert) |
| DEM source | MRDEM 30 m national VRT | OIH Enforced DEM (Ontario) |
| Hydroweight `target_O` | Snapped pour **point** (no area) | Lake **polygon** (has area → `remove_region`) |
| DEM/flow products | Per-`group_id` | One project-wide raster |
| LOI(s) demonstrated | NDVI, NDVI trend, harvest/regen | CANLCC land cover |

Four projects currently run on the engine — all of them one of the two kinds above, never a third:

| Project | Kind | Sites | Terrain | Grouping |
|---|---|---|---|---|
| CELESTE | Stream | 132 | MRDEM (raw, breached per group) | `"hydrobasins"` |
| CAM streams | Stream | 38 | OIH (pre-conditioned) | `"whole_domain"` |
| EMILY_TURKEY | Stream | 177 | OIH (pre-conditioned), **two disjoint tiles** | `"manual_groups"` |
| CAM lakes | Lake | 45 | OIH (pre-conditioned) | `"whole_domain"` |

Each project's own `workflow/<PROJECT>/README.md` has the full rationale, data paths, and one-off corrections specific to it — not duplicated here, since this file documents the shared engine, not any one project's run of it.

::: {.callout-note}
A fifth project, `EMILY_NHN`, was built to delineate the same 177 sites as EMILY_TURKEY via CELESTE's own MRDEM+NHN+HydroBasins approach, as a direct head-to-head comparison against EMILY_TURKEY's OIH+`manual_groups` approach. It was abandoned in favour of EMILY_TURKEY before ever being run — the code remains under `workflow/EMILY_NHN/` as a record, but `output/EMILY_NHN/` was never produced and shouldn't be expected to exist.
:::

### Two-tier processing: groups and sites

Expensive DEM conditioning (breach depressions, flow direction, flow accumulation) is run once per **group** — a hydrologically coherent region sharing a DEM — never once per site. A **site** is a single pour point (or lake); several sites sharing a `group_id` reuse the same conditioned DEM and flow-accumulation raster. In the stream pipeline, `NBE` is one such group: all 8 NBE sites share `cache/CELESTE/NBE/dem.tif`, `flow_pointer.tif`, and `flow_accum.tif`.

How a project's sites get split into groups is itself pluggable (`config$grouping$strategy`, resolved in `engine/01_build_group_manifest.R`):

- **`"hydrobasins"`** — CELESTE. One group per HydroBasins level-6 region a project's sites fall in; used whenever terrain is a single national mosaic (MRDEM) that needs a bounded AOI cropped out of it.
- **`"whole_domain"`** — CAM streams and CAM lakes. One trivial group, no cropping; used whenever one pre-conditioned raster already covers every site.
- **`"manual_groups"`** — EMILY_TURKEY. One group per terrain *source*, auto-detected by which sites each source's raster actually covers. Added specifically because EMILY_TURKEY's 177 sites split across two disjoint OIH tiles with zero overlap (`IntegratedHydrologyNE` / `IntegratedHydrologyNC`) — neither `"whole_domain"` (assumes one raster covers everyone) nor `"hydrobasins"` (assumes one shared national mosaic) fit that.

::: {.columns}

::: {.column width="47%"}
**Stream pipeline**

```
Sites table
   │
   ▼
1. Group manifest
2. Prepare DEM
3. Burn NHN streams (optional)
4. WhiteboxTools conditioning
5. Delineate sites
6. Remove upstream (nesting)
7. Reclip outputs
8. Catchment metrics
9. Hydroweight LOIs
```
:::

::: {.column width="6%"}
:::

::: {.column width="47%"}
**Lake pipeline**

```
Lake sites table
   │
   ▼
1. Match lake polygons
2. Prepare DEM/flow
3. Delineate lakes
4. Remove upstream lakes
5. Reclip outputs
6. Catchment metrics
7. Hydroweight LOIs
```
:::

:::

### Repository layout

```
workflow/
├── R/
│   ├── utils.R                   shared: logging, validation, AOI helpers
│   ├── stream/                   reusable stream/river building blocks —
│   │   │                         used directly by engine/, not a pipeline
│   │   │                         of their own. No numeric prefixes — order
│   │   │                         is set by the caller (the engine).
│   │   ├── group_sites.R
│   │   ├── burn_streams.R
│   │   ├── delineate_sites.R
│   │   └── hydroweight_attributes.R
│   ├── lake/                     reusable lake building blocks, same role
│   │   │                         as stream/ — terrain prep + delineation
│   │   │                         go through engine/, not this folder
│   │   ├── match_lake_polygons.R
│   │   ├── remove_upstream_lakes.R
│   │   └── hydroweight_attributes.R
│   ├── engine/                   modular, input-driven delineation engine —
│   │   │                         standard approach for every project
│   │   ├── 00_resolve_config.R
│   │   ├── 01_build_group_manifest.R   whole_domain / hydrobasins / manual_groups
│   │   ├── 02_prepare_terrain.R
│   │   ├── 03_prepare_streams_burn.R
│   │   ├── 04_delineate_site.R
│   │   ├── prepare_lake_conditioning.R  upfront lake-bisection prevention (opt-in)
│   │   └── 99_rerun_sites
│   ├── remove_upstream.R         shared (stream nesting)
│   ├── reclip_outputs.R          shared (stream + lake)
│   ├── catchment_metrics.R       shared (stream + lake)
│   └── lake_containment.R        shared: lake-bisection detection, reused by
│                                  every project's own fix_lake_bisection.R
├── raster_attributes.R           generic LOI-prep toolbox (project-agnostic)
├── gee_utils.R                   generic Google Earth Engine helpers (project-agnostic)
├── CELESTE/                      CELESTE-specific: runner, LOI prep, README
│   ├── run_celeste.R
│   ├── README.md                 data paths, config rationale, one-off corrections
│   ├── prepare_ndvi.R
│   ├── prepare_ndvi_masked.R     lake-masked NDVI LOI
│   ├── prepare_ndvi_trend.R
│   ├── prepare_harvest_regen.R
│   ├── tidy_outputs.R            long-format plotting-ready tables
│   ├── fix_lake_bisection.R      reactive fallback (see Cross-cutting concepts)
│   └── one_off/                  historical one-off/backfill scripts, kept for the record
├── CAM/                          CAM streams-specific: runner, LOI prep, README
│   ├── run_cam_streams.R
│   ├── README.md
│   ├── prepare_ndvi.R
│   ├── prepare_ndvi_masked.R
│   ├── prepare_ndvi_trend.R
│   ├── prepare_harvest_regen.R
│   ├── upload_catchments_to_gee.R
│   ├── tidy_outputs.R
│   ├── fix_lake_bisection.R
│   └── one_off/
├── EMILY_TURKEY/                 177 sites, OIH terrain (two-tile manual_groups)
│   ├── run_emily_turkey.R
│   ├── README.md                 the two-tile discovery, manual_groups rationale
│   ├── tidy_outputs.R
│   └── one_off/
├── EMILY_NHN/                    abandoned in favour of EMILY_TURKEY, never run —
│   │                             kept as a record, not an active project
│   ├── run_emily_nhn.R
│   └── assign_hydrobasins_groups.R
├── one_off/                      project-agnostic historical one-off/backfill scripts
├── templates/
│   └── run_engine_template.R     blank TODO-annotated skeleton for a new project
└── run_cam_lakes.R                CAM lakes runner

code/
└── reset_workflow.R              cache/output reset helpers

docs/
└── catchment_delineation_guide.qmd   this document

cache/<PROJECT>/<group_id>/  or  cache/<PROJECT>/   group-level rasters (auto-created;
output/<PROJECT>/<site_id>/                          flat cache under whole_domain grouping)
```

Anything specific to one project — its runner, LOI `prepare_*.R` scripts, `fix_lake_bisection.R`, and `one_off/` corrections — lives entirely inside that project's own `workflow/<PROJECT>/` folder, each with a `README.md` carrying the rationale, data paths, and one-off decisions that don't belong in the project-agnostic root `CLAUDE.md`. This is deliberate: it keeps one project's data-prep and correction quirks from ever needing to be reconciled with another's, and it's why this guide documents the shared engine and cross-cutting concepts rather than walking through every project one by one — read a project's own README for that.

---

# Worked example A — Stream delineation (`NBE` group)

`NBE1`–`NBE8` are 8 New Brunswick pour points sharing `group_id = "NBE"`. Every figure in this section reads directly from `cache/CELESTE/NBE/` and `output/CELESTE/NBE*/` — files already produced by the current, engine-based `run_celeste.R` (see the note at the top of this document for how this section's paths changed on 2026-08-31).

## Stage 1 — Group manifest

`build_group_manifest()` uses HydroBasins to build one processing AOI per `group_id`, unioning the basins containing that group's sites. All 9 stages downstream for a given group operate inside this one AOI.

```{r}
#| label: nbe-group-map
#| fig-cap: "NBE pour points on an OpenStreetMap basemap, zoomed to the site cluster. Inset: the full group AOI within New Brunswick."
group_aois <- st_read("cache/group_aois.gpkg", quiet = TRUE)
nbe_aoi <- group_aois |> dplyr::filter(group_id == "NBE")

nbe_pour_points <- list.files("output/CELESTE", pattern = "^NBE[0-9]$", full.names = TRUE) |>
  map(\(d) st_read(file.path(d, "pour_point.gpkg"), quiet = TRUE)) |>
  bind_rows()

# NBE1's catchment is reused as "the example catchment" by several later
# stages, so its extent stays consistent throughout this section.
nbe1_catchment <- st_read("output/CELESTE/NBE1/catchment.gpkg", quiet = TRUE)

zoom_bbox <- st_bbox(st_buffer(nbe_pour_points, 3000))

# maptiles::get_tiles() reprojects into whatever CRS its input sfc uses (here
# EPSG:3979), so the tile raster and coord_sf() stay in the same CRS — plain
# ggspatial::annotation_map_tile() assumes lon/lat and silently mis-locates
# non-4326 data, confirmed while building this figure.
basemap <- maptiles::get_tiles(st_as_sfc(zoom_bbox), provider = "OpenStreetMap", crop = TRUE, zoom = 12)

p_main <- ggplot() +
  tidyterra::geom_spatraster_rgb(data = basemap) +
  geom_sf(data = nbe_aoi, fill = NA, colour = "grey20", linewidth = 0.6) +
  geom_sf(data = nbe_pour_points, aes(colour = site_id), size = 2.6) +
  coord_sf(xlim = c(zoom_bbox["xmin"], zoom_bbox["xmax"]),
           ylim = c(zoom_bbox["ymin"], zoom_bbox["ymax"]), expand = FALSE) +
  scale_colour_manual(values = pal_nbe, name = "Site") +
  theme_cw() +
  labs(title = "NBE pour points",
       subtitle = glue::glue("buffer_m = {nbe_aoi$buffer_m}, burn_streams = {nbe_aoi$burn_streams}, n_sites = {nbe_aoi$n_sites}"))

nb <- rnaturalearth::ne_states(country = "Canada", returnclass = "sf") |>
  dplyr::filter(name == "New Brunswick") |>
  st_transform(3979)

p_inset <- ggplot() +
  geom_sf(data = nb, fill = "grey95", colour = "grey50") +
  geom_sf(data = nbe_aoi, fill = "#D55E00", colour = NA, alpha = 0.85) +
  theme_void() +
  theme(panel.background = element_rect(fill = "white", colour = "grey60"))

p_main + inset_element(p_inset, left = 0.01, bottom = 0.6, right = 0.4, top = 0.99)
```

```{r}
#| label: group-manifest-code
#| eval: false
#| code-fold: false
group_manifest <- build_group_manifest(
  sites            = sites,
  output_dir       = here("output/CELESTE"),
  cache_dir        = here("cache"),
  hydrobasins_dir  = HYDROBASINS_DIR,
  default_buffer_m = 1000
)
```

::: {.callout-warning}
Always inspect group AOIs (`cache/group_aois.gpkg`) before Stage 2. A group AOI that's too small truncates upstream area and produces an incorrect catchment — the fix is to raise `aoi_buffer_m` for that group, clear its cache, and rerun from Stage 2.
:::

## Stage 2 — Prepare DEM

The MRDEM is a GDAL VRT streamed from cloud tiles — `prepare_dem()` crops it to each group AOI and writes `cache/<group_id>/dem.tif`, a rectangular crop (never masked to the AOI polygon — masking would introduce NoData at the boundary that WhiteboxTools reads as depressions, corrupting flow direction near the edge). `cache/CELESTE/NBE/dem.tif` itself is ~100M cells (the full group AOI shown in the Stage 1 inset); the figure below crops it down to the example catchment's extent for display.

```{r}
#| label: nbe-dem
#| fig-cap: "Group DEM cropped to the NBE1 catchment extent, for display — prepare_dem() itself crops to the full group AOI, not this window."
dem_window   <- st_buffer(st_as_sfc(st_bbox(nbe1_catchment)), 1500)
nbe_dem_win  <- rast("cache/CELESTE/NBE/dem.tif") |> terra::crop(vect(dem_window))
nbe_hill_win <- terra::shade(
  terra::terrain(nbe_dem_win, "slope", unit = "radians"),
  terra::terrain(nbe_dem_win, "aspect", unit = "radians")
)

plot_relief(nbe_hill_win, nbe_dem_win, title = "Group DEM near NBE1") +
  geom_sf(data = nbe1_catchment, fill = NA, colour = "white", linewidth = 0.7)
```

```{r}
#| label: prepare-dem-code
#| eval: false
#| code-fold: false
prepare_dem(group_manifest, mrdem_vrt = MRDEM_VRT_PATH, target_crs = 3979)
```

## Stage 3 — Burn NHN streams (optional, per group)

When `burn_streams = TRUE` for a group, NHN flowlines are merged (fixing two known geometry issues — dropping M-coordinates and casting to a single `MULTILINESTRING` type) and burned into the DEM before conditioning, anchoring flow paths to mapped channels. NBE has `burn_streams = TRUE`; its cached flowlines cover the whole group AOI, so the figure below crops to a small window around `NBE1` rather than plotting the full group.

Only `flowlines.gpkg` is persisted to the group cache — a companion `waterbodies.gpkg` used to be written too, but the engine deliberately dropped it: nothing downstream reads a persisted group-level waterbodies file (lake-bisection checking, see Cross-cutting concepts below, does its own on-the-fly NHN fetch instead), so writing one unconditionally would add real runtime to every group's terrain prep for no consumer.

```{r}
#| label: nbe-streams
#| fig-cap: "NHN flowlines burned into the DEM near NBE1 (small window; the full NBE cache holds thousands of flowline features across the group AOI)."
window    <- st_buffer(st_as_sfc(st_bbox(nbe1_catchment)), 2000)
window_bb <- st_bbox(window)

flowlines_win <- st_read("cache/CELESTE/NBE/flowlines.gpkg", quiet = TRUE, wkt_filter = st_as_text(window))

# wkt_filter keeps any feature that *intersects* the window, not just the
# part of it inside the window — a flowline crossing the edge can extend far
# beyond it, so coord_sf() must set the crop explicitly rather than let it
# auto-fit to every layer's full (unclipped) extent.
ggplot() +
  geom_sf(data = window, fill = "grey97", colour = NA) +
  geom_sf(data = flowlines_win, colour = "#1f78b4", linewidth = 0.4) +
  geom_sf(data = nbe1_catchment, fill = NA, colour = "#D55E00", linewidth = 0.6) +
  coord_sf(xlim = c(window_bb["xmin"], window_bb["xmax"]),
           ylim = c(window_bb["ymin"], window_bb["ymax"]), expand = FALSE) +
  theme_cw() +
  labs(title = "NHN network near NBE1", subtitle = "Orange = NBE1 unclipped catchment boundary")
```

```{r}
#| label: nhn-geom-fix
#| eval: false
#| code-fold: false
# Applied internally in burn_streams_into_dem() before WhiteboxTools sees it:
flowlines |>
  sf::st_zm(drop = TRUE, what = "ZM") |>
  sf::st_cast("MULTILINESTRING") |>
  sf::st_write(streams_shp, delete_dsn = TRUE, quiet = TRUE)
```

::: {.callout-tip}
Groups with no matching NHN sheets get an empty `flowlines.gpkg` and a warning, then continue with `burn_streams = FALSE` — a missing NHN download never hard-stops the workflow.
:::

## Stage 4 — WhiteboxTools conditioning

Five steps run per group: breach depressions (least-cost, `dist = 10` cells, `fill = TRUE` to close remaining pits), D8 pointer, D8 flow accumulation (**must** take the pointer, not the DEM, with `pntr = TRUE` — passing the DEM silently re-derives flow direction and bypasses the stream-burned conditioning), stream extraction by threshold, and hillshade.

```{r}
#| label: nbe1-flow-accum
#| fig-cap: "D8 flow pointer (left) and flow accumulation (right) over the NBE1 unclipped catchment. These are Stage 5's per-site outputs (flow_accum.tif/flow_pointer.tif, not the Stage 7 *_clipped versions) — Stage 6/7's clipped versions cover only the much smaller downstream wedge that survives upstream removal (see Stage 6), which would leave most of this extent NA."
fa   <- rast("output/CELESTE/NBE1/flow_accum.tif")
ptr  <- rast("output/CELESTE/NBE1/flow_pointer.tif")
strm <- st_read("output/CELESTE/NBE1/streams.gpkg", quiet = TRUE)

dir_levels <- data.frame(ID = c(1, 2, 4, 8, 16, 32, 64, 128),
                          Direction = c("E", "SE", "S", "SW", "W", "NW", "N", "NE"))
levels(ptr) <- dir_levels

p_ptr <- ggplot() +
  tidyterra::geom_spatraster(data = ptr) +
  scale_fill_manual(values = pal_dir, name = "Flow to", na.translate = FALSE) +
  geom_sf(data = nbe1_catchment, fill = NA, colour = "black", linewidth = 0.5) +
  coord_sf(expand = FALSE) + theme_cw() +
  labs(title = "D8 flow pointer")

p_fa <- ggplot() +
  tidyterra::geom_spatraster(data = log10(fa + 1)) +
  scale_fill_viridis_c(option = "mako", name = "log10(cells)", na.value = NA) +
  geom_sf(data = strm, colour = "white", linewidth = 0.3) +
  geom_sf(data = nbe1_catchment, fill = NA, colour = "#E69F00", linewidth = 0.6) +
  coord_sf(expand = FALSE) + theme_cw() +
  labs(title = "Flow accumulation", subtitle = "White = extracted streams")

p_ptr + p_fa
```

```{r}
#| label: whitebox-code
#| eval: false
#| code-fold: false
whitebox::wbt_breach_depressions_least_cost(dem = dem_burned, output = dem_breached, dist = 10, fill = TRUE)
whitebox::wbt_d8_pointer(dem = dem_breached, output = flow_pointer)
whitebox::wbt_d8_flow_accumulation(input = flow_pointer, output = flow_accum, out_type = "cells", pntr = TRUE)  # <- pntr = TRUE is critical
whitebox::wbt_extract_streams(flow_accum = flow_accum, output = streams, threshold = stream_threshold)
whitebox::wbt_hillshade(dem = dem_breached, output = hillshade)
```

## Stage 5 — Delineate sites

Per site: write the pour point, snap it onto the nearest stream cell (`wbt_jenson_snap_pour_points()`, within `snap_dist`), delineate the watershed from the snapped point and D8 pointer, convert to polygon, then clip every group raster and the NHN flowlines to the result.

```{r}
#| label: nbe1-delineation
#| fig-cap: "NBE1: raw vs. snapped pour point over the full catchment (left), and a close-up of the snap onto the extracted stream network (right)."
pp_raw     <- st_read("output/CELESTE/NBE1/pour_point.gpkg", quiet = TRUE)
pp_snapped <- st_read("output/CELESTE/NBE1/pour_point_snapped.shp", quiet = TRUE)
dem1       <- rast("output/CELESTE/NBE1/dem.tif")
hill1      <- rast("output/CELESTE/NBE1/hillshade.tif")

p_full <- plot_relief(hill1, dem1, title = "Unclipped catchment") +
  geom_sf(data = nbe1_catchment, fill = NA, colour = "white", linewidth = 0.7) +
  geom_sf(data = pp_raw, aes(shape = "raw"), colour = "#D55E00", size = 2.6) +
  geom_sf(data = pp_snapped, aes(shape = "snapped"), colour = "#0072B2", size = 2.6) +
  scale_shape_manual(values = c(raw = 4, snapped = 16), name = NULL)

# Zoomed-in detail of the snap itself, against the (group-extent) streams
# raster it snapped onto — cropped to a small window so this stays fast.
snap_window <- st_buffer(pp_snapped, 250)
streams_tmp <- rast("output/CELESTE/NBE1/streams_tmp.tif") |> terra::crop(vect(snap_window))
snap_bb     <- st_bbox(snap_window)

# White-fill/black-outline marker for the snapped point — a solid blue dot
# (matching pp_snapped's colour everywhere else) all but disappears against
# the blue stream fill at this zoom, so contrast wins over colour-consistency
# for this one panel.
p_zoom <- ggplot() +
  tidyterra::geom_spatraster(data = streams_tmp, show.legend = FALSE) +
  scale_fill_gradient(low = "#1f78b4", high = "#1f78b4", na.value = "grey92") +
  geom_segment(aes(x = st_coordinates(pp_raw)[, 1], y = st_coordinates(pp_raw)[, 2],
                    xend = st_coordinates(pp_snapped)[, 1], yend = st_coordinates(pp_snapped)[, 2]),
               colour = "grey20", linetype = "dashed", linewidth = 0.4,
               arrow = arrow(length = unit(0.15, "cm"))) +
  geom_sf(data = pp_raw, colour = "#D55E00", shape = 4, size = 3, stroke = 1.3) +
  geom_sf(data = pp_snapped, shape = 21, fill = "white", colour = "black", size = 3, stroke = 1.1) +
  coord_sf(xlim = c(snap_bb["xmin"], snap_bb["xmax"]), ylim = c(snap_bb["ymin"], snap_bb["ymax"]), expand = FALSE) +
  theme_cw() +
  theme(axis.text = element_blank(), axis.ticks = element_blank()) +
  labs(title = "Snap detail", x = NULL, y = NULL,
       subtitle = "Blue = stream cells\nOrange x = raw pour point, white/black dot = snapped")

p_full + p_zoom + plot_layout(widths = c(1.3, 1))
```

`wbt_jenson_snap_pour_points()` moves the raw point onto the nearest cell of the **binary streams raster** (not the raw flow accumulation) within `snap_dist` metres — snapping against flow accumulation directly risks jumping to a distant, unrelated high-accumulation cell instead of the intended nearby channel.

```{r}
#| label: correcting-pour-point
#| eval: false
#| code-fold: false
# If a catchment is obviously wrong, the pour point likely snapped to the
# wrong stream cell. Open pour_point_snapped.shp + streams_tmp.tif in QGIS,
# drag the point onto the correct stream cell, then re-run watershed delineation
# only (skips re-snapping, reads the edited shapefile directly). This example
# predates the engine migration; the pre-engine mechanism shown at the time
# (workflow/R/stream/99_rerun_sites's rerun_site_watershed()) has since been
# retired along with the verification script that was its only remaining
# caller (see git history) — every active pipeline, including the current
# engine-based run_celeste.R, now uses workflow/R/engine/99_rerun_sites's
# rerun_engine_site_watershed() instead (same idea, CRS-dynamic):
source(here("workflow/R/engine/99_rerun_sites"))
rerun_engine_site_watershed("NBE1", sites, group_manifest, config, output_dir)
```

## Stage 6 — Remove upstream (nested catchments)

For sites where one pour point sits downstream of another (either within the same group or a shared network), the smaller upstream catchment(s) are erased from the larger one so each site's catchment reports contributing area only, without double-counting area that already belongs to an upstream site. `NBE1` is a strong real example: 10 upstream catchments were erased, taking its area from 236.9 km² down to 17.7 km².

```{r}
#| label: nbe-upstream-removal
#| fig-cap: "Unclipped vs. clipped area for all 8 NBE sites — the gap is area erased because it belongs to an upstream site."
nbe_nesting <- st_read("output/CELESTE/all_catchments_clipped.gpkg", quiet = TRUE) |>
  st_drop_geometry() |>
  dplyr::filter(str_detect(site_id, "^NBE[0-9]$")) |>
  select(site_id, area_km2_before, area_km2_after, n_erased) |>
  pivot_longer(c(area_km2_before, area_km2_after), names_to = "stage", values_to = "area_km2") |>
  mutate(stage = if_else(stage == "area_km2_before", "unclipped", "clipped"))

ggplot(nbe_nesting, aes(x = fct_rev(site_id), y = area_km2, fill = stage)) +
  geom_col(position = "identity", alpha = 0.85) +
  coord_flip() +
  scale_fill_manual(values = c(unclipped = "grey75", clipped = "#0072B2"), name = NULL) +
  theme_cw() +
  labs(title = "NBE: unclipped (grey, full bar) vs. clipped (blue) catchment area",
       x = NULL, y = expression("Area ("*km^2*")"))
```

```{r}
#| label: nbe-nesting-table
#| tbl-cap: "Upstream catchments erased per NBE site."
nbe_nesting_tbl <- st_read("output/CELESTE/all_catchments_clipped.gpkg", quiet = TRUE) |>
  st_drop_geometry() |>
  dplyr::filter(str_detect(site_id, "^NBE[0-9]$")) |>
  arrange(site_id)

gt(nbe_nesting_tbl) |>
  cols_label(site_id = "Site", area_km2_before = "Unclipped (km²)",
             area_km2_after = "Clipped (km²)", n_erased = "Catchments erased") |>
  fmt_number(c(area_km2_before, area_km2_after), decimals = 1) |>
  tab_header(title = "NBE upstream-removal summary")
```

```{r}
#| label: remove-upstream-code
#| eval: false
#| code-fold: false
group_manifest <- remove_upstream_catchments(group_manifest, sites, output_dir)
# -> writes catchment_clipped.gpkg per site, plus all_catchments_clipped.gpkg
```

## Stage 7 — Reclip outputs

Every raster and the flowlines are re-clipped to `catchment_clipped.gpkg`, producing the `*_clipped.tif` / `streams_clipped.gpkg` files used from here on.

```{r}
#| label: nbe1-clipped
#| fig-cap: "NBE1 after clipping — compare against the unclipped panel in Stage 5."
nbe1_clip     <- st_read("output/CELESTE/NBE1/catchment_clipped.gpkg", quiet = TRUE)
dem1_clip     <- rast("output/CELESTE/NBE1/dem_clipped.tif")
hill1_clip    <- terra::shade(terra::terrain(dem1_clip, "slope", unit = "radians"),
                               terra::terrain(dem1_clip, "aspect", unit = "radians"))

plot_relief(hill1_clip, dem1_clip, title = "NBE1 clipped catchment (17.7 km²)") +
  geom_sf(data = nbe1_clip, fill = NA, colour = "white", linewidth = 0.9) +
  geom_sf(data = pp_snapped, colour = "#0072B2", size = 2.8)
```

## Stage 8 — Morphometric metrics

Geometry, areal (shape), and relief parameters follow Shekar & Mathew (2024), computed for both catchment versions.

```{r}
#| label: nbe-metrics-table
#| tbl-cap: "Selected clipped-catchment metrics for the NBE group."
celeste_metrics <- read_csv("output/CELESTE/catchment_metrics.csv", show_col_types = FALSE)

nbe_metrics <- celeste_metrics |>
  dplyr::filter(str_detect(site_id, "^NBE[0-9]$"), version == "clipped") |>
  select(site_id, area_km2, form_factor, circularity_ratio, elongation_ratio,
         basin_relief_m, mean_slope_deg, drainage_density_km_km2, hypsometric_integral) |>
  arrange(site_id)

gt(nbe_metrics) |>
  cols_label(site_id = "Site", area_km2 = "Area (km²)", form_factor = "Form factor",
             circularity_ratio = "Circularity", elongation_ratio = "Elongation",
             basin_relief_m = "Relief (m)", mean_slope_deg = "Mean slope (°)",
             drainage_density_km_km2 = "Drainage density (km/km²)",
             hypsometric_integral = "Hypsometric integral") |>
  fmt_number(c(area_km2, form_factor, circularity_ratio, elongation_ratio, mean_slope_deg,
               drainage_density_km_km2, hypsometric_integral), decimals = 2) |>
  fmt_number(basin_relief_m, decimals = 0) |>
  tab_header(title = "NBE clipped-catchment morphometrics")
```

```{r}
#| label: nbe-area-vs-relief
#| fig-cap: "Basin relief vs. area for the NBE group (clipped catchments)."
ggplot(nbe_metrics, aes(area_km2, basin_relief_m, colour = site_id, label = site_id)) +
  geom_point(size = 3) +
  ggrepel::geom_text_repel(show.legend = FALSE, size = 3.2) +
  scale_colour_manual(values = pal_nbe, name = "Site") +
  scale_x_log10() +
  theme_cw() +
  labs(title = "NBE catchments: relief vs. area", x = expression("Area, log scale ("*km^2*")"), y = "Basin relief (m)")
```

Two of the metrics above are easiest to understand as pictures rather than formulas — both illustrated below directly from NBE1's real geometry and DEM.

```{r}
#| label: nbe1-shape-diagram
#| fig-cap: "NBE1 (clipped catchment), illustrating two areal metrics: the red diagonal is basin length Lb (diagonal of the axis-aligned bounding box, used by form_factor = A/Lb^2 and elongation_ratio); the dashed blue circle has the same area as the catchment, the reference used by circularity_ratio = A / area of that circle."
bb <- st_bbox(nbe1_clip)
diag_line <- st_sfc(
  st_linestring(matrix(c(bb["xmin"], bb["ymin"], bb["xmax"], bb["ymax"]), ncol = 2, byrow = TRUE)),
  crs = st_crs(nbe1_clip)
)
bbox_poly <- st_as_sfc(bb)

# st_point_on_surface(), not st_centroid() — a centroid can land outside a
# concave polygon (see CLAUDE.md's terra/hydroweight quirks); doesn't matter
# for this illustration, but it's the safe habit either way.
ctr     <- st_point_on_surface(nbe1_clip)
area_m2 <- as.numeric(st_area(nbe1_clip))
circle  <- st_buffer(ctr, sqrt(area_m2 / pi))

p_shape <- ggplot() +
  geom_sf(data = nbe1_clip, fill = "grey90", colour = "grey30") +
  geom_sf(data = bbox_poly, fill = NA, colour = "grey60", linetype = "dotted") +
  geom_sf(data = diag_line, colour = "#D55E00", linewidth = 0.9) +
  geom_sf(data = circle, fill = NA, colour = "#0072B2", linewidth = 0.9, linetype = "dashed") +
  theme_cw() +
  labs(title = "Shape metrics")

# Hypsometric curve — relative area (a/A) vs. relative height (h/H), computed
# directly from the clipped DEM's own elevation distribution. Its area under
# the curve approximates the table's hypsometric_integral for NBE1.
elev <- terra::values(dem1_clip, na.rm = TRUE) |> as.numeric() |> sort(decreasing = TRUE)
hyps <- tibble(
  rel_area   = seq_along(elev) / length(elev),
  rel_height = (elev - min(elev)) / (max(elev) - min(elev))
)

p_hyps <- ggplot(hyps, aes(rel_area, rel_height)) +
  geom_ribbon(aes(ymin = 0, ymax = rel_height), fill = "#0072B2", alpha = 0.15) +
  geom_line(colour = "#0072B2", linewidth = 0.9) +
  theme_cw() +
  labs(title = "Hypsometric curve", x = "Relative area (a/A)", y = "Relative height (h/H)")

p_shape + p_hyps
```

Reference: Shekar, P.R. & Mathew, A. (2024). Morphometric analysis of watersheds: a comprehensive review of data sources, quality, and geospatial techniques. *Watershed Ecology and the Environment*, 6, 13–25.

## Stage 9 — Hydroweight distance-weighted attributes

`target_O` for the stream pipeline is the snapped **pour point** (`output/<site_id>/pour_point.gpkg`) — a point has no `remove_region`, unlike the lake case below. DEM/flow accumulation are resolved per site from that site's own group cache, since stream groups (NIP/TUR/KEN in Ontario, NBE/COC in New Brunswick, MOR in PEI) span independent HydroBasins regions. Each of seven weighting schemes distance-weights every LOI value by its position relative to the pour point:

```{r}
#| label: weighting-schemes-table
#| tbl-cap: "hydroweight weighting schemes."
schemes <- tribble(
  ~Scheme, ~Description,
  "lumped",  "Unweighted areal mean/proportion — every cell counted equally",
  "iEucO",   "Inverse Euclidean distance to the outflow (O) target",
  "iFLO",    "Inverse flow-path distance to the outflow, along the D8 network",
  "HAiFLO",  "iFLO further weighted by hydrologically-active area (flow accumulation)",
  "iEucS",   "Inverse Euclidean distance to the nearest stream cell",
  "iFLS",    "Inverse flow-path distance to the nearest stream cell",
  "HAiFLS",  "iFLS further weighted by hydrologically-active area"
)
gt(schemes)
```

Six LOIs are now prepared for the stream pipeline (`workflow/CELESTE/prepare_*.R`, mirrored under `workflow/CAM/`): continuous NDVI (annual mosaic, 1984–2025) and a per-pixel NDVI Theil-Sen slope/Mann-Kendall trend (one raster per group, ~29 min one-time), each with a lake-masked sibling (`ndvi_masked` / `ndvi_trend_masked`, via `prepare_ndvi_masked.R`) — masked because a lake pixel's NDVI isn't a terrestrial-vegetation reading and otherwise drags a catchment's summarized NDVI down in proportion to how much open water it contains; kept as a *separate* named LOI alongside the unmasked one, never applied in place, so both stay independently inspectable; a categorical harvest/regen layer (Ontario + New Brunswick sources, 2002–2024, one band per year plus a combined band); and CANLCC land cover (added to the stream pipeline after first being demonstrated below for lakes). The two demonstrated below (NDVI trend, harvest/regen) are enough to show how the mechanics — LOI descriptor shapes, weighting schemes — work; the newer ones follow the same pattern.

```{r}
#| label: nbe-hydroweight
#| fig-cap: "NBE: long-term NDVI trend (Theil-Sen slope, unweighted zonal mean) vs. proportion of catchment ever harvested 2002-2024 (areal/lumped), clipped catchments."
celeste_hw <- read_csv("output/CELESTE/CELESTE_hydroweight.csv", show_col_types = FALSE)

nbe_hw <- celeste_hw |>
  dplyr::filter(str_detect(site, "^NBE[0-9]$"), version == "clipped") |>
  select(site, ndvi_trend_slope_mean, harvest_regen_combined_harvest_lumped_prop) |>
  arrange(site)

p1 <- ggplot(nbe_hw, aes(fct_rev(site), ndvi_trend_slope_mean, fill = site)) +
  geom_col(show.legend = FALSE) +
  geom_hline(yintercept = 0, linewidth = 0.3) +
  coord_flip() +
  scale_fill_manual(values = pal_nbe) +
  theme_cw() +
  labs(title = "NDVI trend (Sen's slope/yr)", x = NULL, y = "Slope")

p2 <- ggplot(nbe_hw, aes(fct_rev(site), harvest_regen_combined_harvest_lumped_prop, fill = site)) +
  geom_col(show.legend = FALSE) +
  coord_flip() +
  scale_y_continuous(labels = percent_format()) +
  scale_fill_manual(values = pal_nbe) +
  theme_cw() +
  labs(title = "Ever harvested 2002-2024", x = NULL, y = "% of catchment")

p1 + p2
```

The 7 schemes above aren't interchangeable — they can give visibly different answers for the same site and the same LOI, depending on where within the catchment the disturbed area actually sits relative to the outflow:

```{r}
#| label: nbe-schemes-compare
#| fig-cap: "How the choice of weighting scheme changes the answer: proportion of catchment classed as harvested (2002-2024), per NBE site, under all 7 schemes."
scheme_labels <- c(lumped = "lumped", ieuco = "iEucO", iflo = "iFLO", haiflo = "HAiFLO",
                    ieucs = "iEucS", ifls = "iFLS", haifls = "HAiFLS")

nbe_schemes <- celeste_hw |>
  dplyr::filter(str_detect(site, "^NBE[0-9]$"), version == "clipped") |>
  select(site, matches(glue::glue("^harvest_regen_combined_harvest_({paste(names(scheme_labels), collapse = '|')})_prop$"))) |>
  pivot_longer(-site, names_to = "scheme", values_to = "prop") |>
  mutate(
    scheme = str_remove(scheme, "^harvest_regen_combined_harvest_") |> str_remove("_prop$"),
    scheme = factor(scheme_labels[scheme], levels = scheme_labels)
  )

ggplot(nbe_schemes, aes(scheme, prop, group = site, colour = site)) +
  geom_line(linewidth = 0.6, alpha = 0.8) +
  geom_point(size = 1.8) +
  scale_colour_manual(values = pal_nbe, name = "Site") +
  scale_y_continuous(labels = percent_format()) +
  theme_cw() +
  theme(axis.text.x = element_text(angle = 30, hjust = 1)) +
  labs(title = "Harvested proportion under each weighting scheme", x = NULL, y = "% of catchment")
```

```{r}
#| label: nbe-interactive-map
#| cache: false
#| fig-cap: "Interactive: NBE clipped catchments and pour points (HTML only — a static overview replaces this in the PDF)."
nbe_catchments_clip <- st_read("output/CELESTE/all_catchments_clipped.gpkg", quiet = TRUE) |>
  dplyr::filter(str_detect(site_id, "^NBE[0-9]$"))

if (is_html) {
  suppressPackageStartupMessages(library(tmap))
  tmap_mode("view")
  tm_shape(nbe_catchments_clip) +
    tm_polygons(fill = "site_id", fill.scale = tm_scale(values = pal_nbe), fill_alpha = 0.5,
                fill.legend = tm_legend(title = "Site")) +
    tm_shape(nbe_pour_points) +
    tm_dots(size = 0.6, fill = "black")
} else {
  ggplot() +
    geom_sf(data = nbe_catchments_clip, aes(fill = site_id), colour = "white", alpha = 0.75) +
    geom_sf(data = nbe_pour_points, size = 1.6) +
    scale_fill_manual(values = pal_nbe, name = "Site") +
    theme_cw() +
    labs(title = "NBE clipped catchments overview")
}
```

```{r}
#| label: hydroweight-code-stream
#| eval: false
#| code-fold: false
hw_results <- calculate_hydroweight_attributes_stream(
  sites          = sites,
  group_manifest = group_manifest,
  output_dir     = output_dir,
  cache_dir      = cache_dir,
  loi_layers     = list(
    ndvi          = list(name = "ndvi", type = "continuous", path_lazy = prepare_ndvi_vrt()),
    ndvi_trend    = list(name = "ndvi_trend", type = "continuous", path_template = "…/{group_id}_ndvi_trend.tif"),
    harvest_regen = list(name = "harvest_regen", type = "categorical", path_lazy = "…", class_levels = harvest_regen_levels)
  )
)
```

---

# Worked example B — Lake delineation (`Clearwater`, `Swan`, `Tilton`, `Lohi`)

These 4 lakes share one project-wide OIH DEM/flow-accumulation raster (`cache/CAM/`) rather than a per-group one — the lake pipeline doesn't span disjoint HydroBasins regions the way the stream pipeline's groups do, so there's no equivalent of the group manifest here.

## Stage 1 — Match lake polygons

`match_lake_polygons()` matches each lake site to an OHN waterbody polygon via a three-pass strategy: spatial buffer intersection, fuzzy name matching (Jaro-Winkler against `OFFICIAL_N`), then manual `OGF_ID` lookup for confirmed hard cases. The matched polygon is rasterized (buffered 30 m) to the full D8 extent as the pour-point target for `wbt_watershed()` — the map below reconstructs each lake's footprint from that rasterized pour-point target, over an OpenStreetMap basemap for geographic context.

```{r}
#| label: cam-lakes-map
#| cache: false
#| fig-cap: "Interactive: the 4 example lakes on an OpenStreetMap basemap (HTML only — small-multiple outlines replace this in the PDF). Basemap tiles are fetched live by the viewer's browser, not embedded at render time."
cam_lakes <- c("Clearwater", "Swan", "Tilton", "Lohi")

lake_outline <- function(lake) {
  r <- rast(glue::glue("output/CAM/lake_delineation/{lake}/lake_pourpoint.tif")) |> terra::trim()
  terra::as.polygons(r, dissolve = TRUE) |> st_as_sf() |> mutate(site_id = lake)
}
cam_lake_polys <- map(cam_lakes, lake_outline) |> bind_rows()

if (is_html) {
  suppressPackageStartupMessages(library(tmap))
  tmap_mode("view")
  tm_basemap("OpenStreetMap") +
    tm_shape(cam_lake_polys) +
    tm_polygons(fill = "site_id", fill.scale = tm_scale(values = pal_lakes),
                fill.legend = tm_legend(title = "Lake"))
} else {
  # facet_wrap(scales = "free") isn't supported with coord_sf(), and these 4
  # lakes are geographically far apart — so each gets its own small map (own
  # extent) combined with patchwork, rather than one shared-extent facet grid.
  lake_maps <- map(cam_lakes, \(lake) {
    ggplot(dplyr::filter(cam_lake_polys, site_id == lake)) +
      geom_sf(fill = pal_lakes[[lake]], colour = "grey20") +
      theme_cw() +
      theme(axis.text = element_blank(), legend.position = "none") +
      labs(title = lake)
  })
  wrap_plots(lake_maps, nrow = 1) +
    plot_annotation(title = "Matched lake polygons (from rasterized pour-point target)")
}
```

```{r}
#| label: match-lakes-code
#| eval: false
#| code-fold: false
lake_polys <- match_lake_polygons(
  sites            = sites,
  lakes_path       = OIH_LAKES_PATH,
  buffer_m         = 10,
  name_dist_max    = 0.15,
  name_search_km   = 40
)
```

## Stage 2 — Prepare OIH DEM and flow products

One project-wide DEM and D8 flow-accumulation raster are prepared for the whole of the OIH-covered Ontario domain (not per group, unlike the stream pipeline) — `cache/CAM/dem.tif`, `d8_pntr.tif`, `flow_accum.tif`.

```{r}
#| label: clearwater-relief
#| fig-cap: "Clearwater lake catchment: relief, from its own clipped rasters. The lake is shown as an outline (not filled) so the DEM underneath it stays visible."
cw_dem   <- rast("output/CAM/lake_delineation/Clearwater/dem.tif")
cw_hill  <- terra::shade(terra::terrain(cw_dem, "slope", unit = "radians"),
                          terra::terrain(cw_dem, "aspect", unit = "radians"))
cw_catch <- st_read("output/CAM/lake_delineation/Clearwater/catchment.gpkg", quiet = TRUE)

plot_relief(cw_hill, cw_dem, title = "Clearwater catchment") +
  geom_sf(data = cw_catch, fill = NA, colour = "white", linewidth = 0.8) +
  geom_sf(data = st_as_sf(cam_lake_polys[cam_lake_polys$site_id == "Clearwater", ]),
          fill = NA, colour = "#0072B2", linewidth = 1)
```

## Stage 3 — Delineate lake catchments

The matched (buffered) lake polygon is rasterized as the pour-point target and passed to `wbt_watershed()` — analogous to the snapped point in the stream pipeline, but an area rather than a single cell.

```{r}
#| label: cam-4lakes-catchments
#| fig-cap: "Unclipped catchments for the 4 example lakes. Each is outlined in black so overlapping/adjacent boundaries stay legible even where fills are similar."
cam_catchments <- st_read("output/CAM/lake_delineation/all_catchments.gpkg", quiet = TRUE) |>
  dplyr::filter(site_id %in% cam_lakes)

ggplot(cam_catchments) +
  geom_sf(aes(fill = site_id), colour = "black", linewidth = 0.4, alpha = 0.85) +
  geom_sf(data = cam_lake_polys, fill = "grey15", colour = NA) +
  scale_fill_manual(values = pal_lakes, name = "Lake") +
  theme_cw() +
  labs(title = "Clearwater, Swan, Tilton, Lohi: unclipped catchments", subtitle = "Dark fill = lake footprint")
```

```{r}
#| label: delineate-lakes-code
#| eval: false
#| code-fold: false
results <- delineate_lake_catchments(
  sites      = sites,
  lake_polys = lake_polys,
  cache_dir  = cache_dir,
  output_dir = output_dir
)
```

## Stage 4 — Remove upstream lakes

Same nesting concept as Stage 6 for streams, applied to lake catchments. Among these 4 lakes, `Swan`'s clipped area is identical to its unclipped area (no upstream lake to erase), while `Lohi` loses most of its area to upstream removal — the interactive map below lets you toggle the unclipped and clipped layers to see exactly which lakes sit upstream of which.

```{r}
#| label: cam-4lakes-nesting
#| fig-cap: "Unclipped vs. clipped area for the 4 example lakes (from catchment_metrics.csv)."
cam_metrics <- read_csv("output/CAM/lake_delineation/catchment_metrics.csv", show_col_types = FALSE)

cam_nesting <- cam_metrics |>
  dplyr::filter(site_id %in% cam_lakes) |>
  select(site_id, version, area_km2) |>
  pivot_wider(names_from = version, values_from = area_km2)

gt(cam_nesting) |>
  cols_label(site_id = "Lake", unclipped = "Unclipped (km²)", clipped = "Clipped (km²)") |>
  fmt_number(c(unclipped, clipped), decimals = 2) |>
  tab_header(title = "CAM upstream-removal effect, 4 example lakes")
```

```{r}
#| label: cam-nesting-interactive
#| cache: false
#| fig-cap: "Interactive: unclipped catchments (pale, full extent) vs. clipped catchments (solid, black outline) for the 4 example lakes — use the layer control to toggle each on/off and see how the smaller lakes nest inside the larger ones. HTML only."
cam_catchments_unclip <- st_read("output/CAM/lake_delineation/all_catchments.gpkg", quiet = TRUE) |>
  dplyr::filter(site_id %in% cam_lakes)
cam_catchments_clip4 <- st_read("output/CAM/lake_delineation/all_catchments_clipped.gpkg", quiet = TRUE) |>
  dplyr::filter(site_id %in% cam_lakes)

if (is_html) {
  suppressPackageStartupMessages(library(tmap))
  tmap_mode("view")
  tm_shape(cam_catchments_unclip) +
    tm_polygons(fill = "site_id", fill.scale = tm_scale(values = pal_lakes), fill_alpha = 0.25,
                fill.legend = tm_legend(show = FALSE), group = "Unclipped catchments") +
    tm_shape(cam_catchments_clip4) +
    tm_polygons(fill = "site_id", fill.scale = tm_scale(values = pal_lakes), fill_alpha = 0.7,
                col = "black", lwd = 1, fill.legend = tm_legend(title = "Lake"), group = "Clipped catchments") +
    tm_shape(cam_lake_polys) +
    tm_polygons(fill = "grey15", fill.legend = tm_legend(show = FALSE), group = "Lakes")
} else {
  ggplot() +
    geom_sf(data = cam_catchments_unclip, aes(fill = site_id), alpha = 0.25, colour = NA) +
    geom_sf(data = cam_catchments_clip4, aes(fill = site_id), colour = "black", alpha = 0.7) +
    geom_sf(data = cam_lake_polys, fill = "grey15", colour = NA) +
    scale_fill_manual(values = pal_lakes, name = "Lake") +
    theme_cw() +
    labs(title = "Unclipped (pale) vs. clipped (solid) catchments")
}
```

```{r}
#| label: remove-upstream-lake-code
#| eval: false
#| code-fold: false
lake_polys <- remove_upstream_lake_catchments(lake_polys, sites, cache_dir, output_dir)
```

## Stage 5 — Reclip outputs

Same `reclip_outputs.R` module as the stream pipeline — it works from either a `group_manifest` (stream) or a `cache_dir` (lake), producing the same `*_clipped.tif` / `streams_clipped.gpkg` convention either way.

## Stage 6 — Morphometric metrics

The same `catchment_metrics.R` module also reports lake-specific morphometric parameters (fetch, shoreline development, lake area %, catchment-area:lake-area ratio) alongside the standard geometry/areal/relief set, when a lake polygon is present.

```{r}
#| label: cam-metrics-table
#| tbl-cap: "Selected clipped-catchment metrics, 4 example lakes."
cam_metrics_tbl <- cam_metrics |>
  dplyr::filter(site_id %in% cam_lakes, version == "clipped") |>
  select(site_id, area_km2, lake_area_ha, ca_la_ratio, shoreline_development,
         mean_slope_deg, drainage_density_km_km2) |>
  arrange(site_id)

gt(cam_metrics_tbl) |>
  cols_label(site_id = "Lake", area_km2 = "Catchment area (km²)", lake_area_ha = "Lake area (ha)",
             ca_la_ratio = "CA:LA ratio", shoreline_development = "Shoreline dev.",
             mean_slope_deg = "Mean slope (°)", drainage_density_km_km2 = "Drainage density") |>
  fmt_number(everything(), decimals = 2) |>
  tab_header(title = "CAM clipped-catchment morphometrics")
```

These 4 lakes were picked as examples because they genuinely span a wide range on most of these metrics — from `Swan`'s tiny, near-1:1 catchment to `Lohi`'s much larger, higher-ratio one. As in Stage 8 for the stream pipeline, that variation is easier to read as pictures than as table rows: a shape diagram per lake, and their hypsometric curves overlaid.

```{r}
#| label: cam-shape-diagrams
#| fig-cap: "Shape metrics for each of the 4 example lake catchments: the coloured diagonal is basin length Lb; the dashed circle is the equal-area circle used by circularity_ratio. Each lake gets its own extent (small multiples, as in Stage 1) since they're geographically far apart."
cam_shape_diagram <- function(lake) {
  catch <- st_read(glue::glue("output/CAM/lake_delineation/{lake}/catchment_clipped.gpkg"), quiet = TRUE)

  bb <- st_bbox(catch)
  diag_line <- st_sfc(
    st_linestring(matrix(c(bb["xmin"], bb["ymin"], bb["xmax"], bb["ymax"]), ncol = 2, byrow = TRUE)),
    crs = st_crs(catch)
  )
  bbox_poly <- st_as_sfc(bb)

  ctr     <- st_point_on_surface(catch)
  area_m2 <- as.numeric(st_area(catch))
  circle  <- st_buffer(ctr, sqrt(area_m2 / pi))

  ggplot() +
    geom_sf(data = catch, fill = "grey90", colour = "grey30") +
    geom_sf(data = bbox_poly, fill = NA, colour = "grey60", linetype = "dotted") +
    geom_sf(data = diag_line, colour = pal_lakes[[lake]], linewidth = 1) +
    geom_sf(data = circle, fill = NA, colour = "black", linewidth = 0.7, linetype = "dashed") +
    theme_cw() +
    theme(axis.text = element_blank()) +
    labs(title = lake)
}

wrap_plots(map(cam_lakes, cam_shape_diagram), nrow = 1)
```

```{r}
#| label: cam-hypsometric
#| fig-cap: "Hypsometric curves for the 4 example lake catchments, computed directly from each lake's dem_clipped.tif."
cam_hypsometric <- function(lake) {
  dem  <- rast(glue::glue("output/CAM/lake_delineation/{lake}/dem_clipped.tif"))
  elev <- terra::values(dem, na.rm = TRUE) |> as.numeric() |> sort(decreasing = TRUE)
  tibble(
    site_id    = lake,
    rel_area   = seq_along(elev) / length(elev),
    rel_height = (elev - min(elev)) / (max(elev) - min(elev))
  )
}

cam_hyps <- map(cam_lakes, cam_hypsometric) |> bind_rows()

ggplot(cam_hyps, aes(rel_area, rel_height, colour = site_id)) +
  geom_line(linewidth = 0.9) +
  scale_colour_manual(values = pal_lakes, name = "Lake") +
  theme_cw() +
  labs(title = "Hypsometric curves", x = "Relative area (a/A)", y = "Relative height (h/H)")
```

## Stage 7 — Hydroweight distance-weighted attributes

`target_O` here is the **lake polygon** itself (not a point), so `remove_region` excises the lake's own footprint from LOI summaries — a stream pour point has no area to exclude. CAM's demonstrated LOI is CANLCC 2020 land cover, reprojected once to EPSG:3161 and cached (`cache/CAM/hydroweight_loi/canlcc_reprojected.tif`).

```{r}
#| label: cam-hydroweight
#| fig-cap: "Land cover composition of the 4 example lake catchments (lumped/areal weighting)."
cam_hw <- read_csv("output/CAM/lake_delineation/CAM_hydroweight.csv", show_col_types = FALSE)

cam_hw_long <- cam_hw |>
  dplyr::filter(site %in% cam_lakes) |>
  select(site, ends_with("_lumped_prop")) |>
  pivot_longer(-site, names_to = "class", values_to = "prop") |>
  mutate(class = str_remove(class, "^canlcc_") |> str_remove("_lumped_prop$") |> str_replace_all("_", " ") |> str_to_title())

ggplot(cam_hw_long, aes(x = site, y = prop, fill = class)) +
  geom_col(position = "stack") +
  scale_y_continuous(labels = percent_format()) +
  # 10 CANLCC classes — beyond the 8-colour Okabe-Ito set used elsewhere for
  # site/lake identity, so a larger qualitative palette is used here instead.
  scale_fill_brewer(palette = "Paired", name = "CANLCC class") +
  theme_cw() +
  labs(title = "CANLCC land cover composition (lumped)", x = NULL, y = "% of catchment")
```

```{r}
#| label: cam-interactive-map
#| cache: false
#| fig-cap: "Interactive: the 4 example lake catchments with the rasterized CANLCC land cover layer underneath (HTML only — a static overview replaces this in the PDF)."
cam_catchments_clip <- st_read("output/CAM/lake_delineation/all_catchments_clipped.gpkg", quiet = TRUE) |>
  dplyr::filter(site_id %in% cam_lakes)

if (is_html) {
  suppressPackageStartupMessages(library(tmap))

  # The full CANLCC raster is huge (61225 x 62182 cells); cropping first —
  # rather than reading it whole — is what keeps this fast (same pattern
  # raster_attributes.R uses for any path_lazy LOI source).
  canlcc_levels_lookup <- data.frame(
    ID    = c(1L, 2L, 5L, 6L, 8L, 10L, 11L, 12L, 13L, 14L, 15L, 16L, 17L, 18L, 19L),
    Class = c("needleleaf_forest", "taiga_needleleaf_forest", "broadleaf_deciduous_forest",
              "mixed_forest", "temperate_shrubland", "temperate_grassland", "shrubland_lichen_moss",
              "grassland_lichen_moss", "barren_lichen_moss", "wetland", "cropland", "barren",
              "urban", "water", "snow_ice")
  )
  canlcc_bb   <- st_bbox(st_buffer(cam_catchments_clip, 500))
  canlcc_crop <- rast("cache/CAM/hydroweight_loi/canlcc_reprojected.tif") |>
    terra::crop(terra::ext(canlcc_bb["xmin"], canlcc_bb["xmax"], canlcc_bb["ymin"], canlcc_bb["ymax"]))
  levels(canlcc_crop) <- canlcc_levels_lookup

  tmap_mode("view")
  tm_shape(canlcc_crop) +
    tm_raster(col.legend = tm_legend(title = "Land cover")) +
    tm_shape(cam_catchments_clip) +
    tm_polygons(fill = "site_id", fill.scale = tm_scale(values = pal_lakes), fill_alpha = 0.35,
                col = "black", lwd = 1.5, fill.legend = tm_legend(title = "Lake"))
} else {
  ggplot() +
    geom_sf(data = cam_catchments_clip, aes(fill = site_id), colour = "black", alpha = 0.75) +
    geom_sf(data = cam_lake_polys, fill = "grey15", colour = NA) +
    scale_fill_manual(values = pal_lakes, name = "Lake") +
    theme_cw() +
    labs(title = "CAM clipped catchments overview")
}
```

```{r}
#| label: hydroweight-code-lake
#| eval: false
#| code-fold: false
hw_results <- calculate_hydroweight_attributes(
  sites      = sites,
  lake_polys = lake_polys,
  cache_dir  = cache_dir,
  output_dir = output_dir,
  loi_layers = list(
    canlcc = list(name = "canlcc", type = "categorical", path = CANLCC_PATH, class_levels = canlcc_levels)
  )
)
```

---

# Cross-cutting concepts

## LOI layer descriptor shapes

`loi_layers[[i]]` supports exactly one of three source shapes, chosen per LOI based on how the source data is distributed:

| Shape | When to use | Example |
|---|---|---|
| `path` | One raster already covering the full site extent | CANLCC (lake pipeline) |
| `path_template` | One pre-clipped raster per site, `"{site_id}"` placeholder | — |
| `path_lazy` | Shared source too large to reproject/cache whole (mosaic VRT, or one raster per `"{group_id}"`); cropped to each site's catchment *before* being reprojected or written to disk | NDVI mosaic VRT, NDVI trend, harvest/regen (stream pipeline) |

## Categorical LOI quirks (found the hard way)

A handful of `terra`/`hydroweight` behaviours are easy to get wrong silently when building a multi-band categorical LOI like harvest/regen. All were confirmed by direct testing while building that layer (24 bands: 23 years + combined):

- `terra::levels(r) <- df` on a **multi-layer** SpatRaster only applies to layer 1 — layers 2+ silently keep default auto-generated labels. Build each band as an independent single-layer raster, set its levels, then combine via `terra::rast(list(...))`; never mutate a multi-layer factor object in place afterwards.
- An exactly-3-band Byte GeoTIFF gets GDAL's automatic RGB interpretation on write, silently renaming bands to "red"/"green"/"blue" on every subsequent read — use `gdal = "PHOTOMETRIC=MINISBLACK"`.
- A class covering 100% of a catchment produces an id-less, degenerate column name — detected and handled directly (its distance-weighted proportion is provably 1.0 under any scheme) rather than passed through `hydroweight_attributes()`.
- A class with zero occurrences (but not the *only* class) is omitted from that row's output entirely, not reported as 0 — becomes a misleading `NA` after binding multiple sites unless explicitly back-filled.
- `sf::st_read()`'s `wkt_filter` is pushed down to OGR *before* R-side geometry cleanup, so a source with curved geometries can silently drop or entirely miss features depending on the source — safer to read the full layer, clean geometry, then filter spatially in R.

## Caching and incremental reruns

Every stage checks `cache_exists(path)` (file exists **and** is non-empty) before writing, so a rerun after a crash or a newly-added site safely skips completed work — only the new site triggers Stage 5+ (delineation onward); the group's DEM conditioning from Stages 2–4 is reused as-is.

```{r}
#| label: cache-exists-code
#| eval: false
#| code-fold: false
cache_exists <- function(path) fs::file_exists(path) && fs::file_size(path) > 0
```

To force specific parts to rerun, `code/reset_workflow.R` provides:

```{r}
#| label: reset-code
#| eval: false
#| code-fold: false
reset_site("NBE1")            # one site's output/ folder only
reset_group("NBE", sites)     # one group's cache + all its sites' outputs
reset_sites(sites)            # every site's outputs, keep the group cache
reset_all()                   # everything — cache and outputs
```

## Correcting a misaligned pour point / lake target

| Pipeline | Symptom | Fix |
|---|---|---|
| Stream | Catchment too small / wrong tributary | Edit `output/<site_id>/pour_point_snapped.shp` in QGIS against `streams_tmp.tif`, then `rerun_engine_site_watershed()` (`workflow/R/engine/99_rerun_sites`) — reads the edited shapefile directly, skips re-snapping |
| Lake | Catchment misses part of the lake, or grabs a neighbouring one | Edit that site's `lake_pourpoint.tif` in QGIS, then `rerun_engine_lake_site_watershed()` (same file) — re-runs watershed from the edited raster without regenerating it from the polygon match |

## Lake bisection: prevention and correction

A catchment that should flow *through* a lake can instead get cut in two by it — the D8 pointer treats the lake's flat surface as ambiguous and routes flow around rather than across, splitting what should be one contiguous catchment into disconnected pieces. Two mechanisms address this, at different points in the pipeline:

- **Upfront lake conditioning (prevention, opt-in per project — `config$lake_conditioning`, `engine/prepare_lake_conditioning.R`).** Flattens known lakes into the DEM *before* D8 derivation, so every site in a group is delineated from one internally-consistent, lake-aware flow field to begin with. The lake search is scoped to a buffer around a group's own sites (not its full processing AOI) — confirmed directly on CELESTE's KEN group that the AOI is ~22x larger than the sites' own bounding box, and searching it took ~68 minutes for one group versus a small fraction of that scoped to the sites. CELESTE runs with this on; verified on 2026-08-29 that it resolved 3/3 originally-bisected KEN sites and took NIP+TUR from 31 bisected (site, lake) pairs down to 2 (both at a documented noise-band edge).
- **Reactive detection and correction (fallback, per-project `fix_lake_bisection.R`, both CELESTE's and CAM's built on the shared, project-agnostic `workflow/R/lake_containment.R`).** For any bisection that gets through anyway (or for a project that hasn't opted into upfront conditioning), scans finished catchments against a lake source, reports the affected `(site, lake)` pairs, and re-delineates just those sites with the lake explicitly conditioned into their group's terrain. CELESTE's and CAM's versions differ only in the lake source they read (NHN per-HydroBasins-group vs. one province-wide OIH file) and in which terrain tier they're patching — both call into the same shared containment-detection logic rather than duplicating it.

Prevention is the primary defense where it's wired in; the reactive fix exists because it predates upfront conditioning and because not every project has opted in.

---

# Troubleshooting

**Interior pit cells warning from WhiteboxTools.** The DEM was masked to the AOI polygon before conditioning (introducing NoData that reads as depressions). `prepare_dem()` crops to the rectangular extent only by design — check nothing external re-masked the DEM before this stage.

**Flow accumulation max value implausibly small.** `wbt_d8_flow_accumulation()` was passed the DEM instead of the flow pointer, or `pntr = TRUE` was omitted — verify both if calling WhiteboxTools functions directly outside the module.

**NHN sheet not found for a group.** The corresponding GDB hasn't been downloaded. The workflow degrades gracefully (empty flowlines, `burn_streams` effectively off for that group) rather than erroring — download the sheet from GeoBase NHN and place it in `nhn_dir` if burning is actually needed there.

**A stream group's LOI mosaic reads as 0 everywhere instead of NA.** `build_mosaic_vrt()` always sets `-vrtnodata` for exactly this reason — without it, any area not covered by *any* input tile reads as `0`, indistinguishable from real data. This was confirmed directly: before the fix, the NBE group (no NDVI tile coverage at the time) silently produced "NDVI = 0" for every NBE catchment instead of failing the "all-NA after crop/mask" check that would have excluded it correctly.

**GEOS/invalid-geometry errors from HydroBasins or waterbody sources.** `st_make_valid()` is applied inside `build_group_aoi()` (stream) and before rasterization in `rasterize_competing_classes()`. If the error persists it's likely a different geometry op — check the traceback for exactly which call raised it.

**A catchment is split into disconnected pieces around a lake.** Lake bisection — see "Lake bisection: prevention and correction" above. If the project has `lake_conditioning` configured, check it's actually enabled for the affected group; otherwise run the project's `fix_lake_bisection.R` to detect and correct it reactively.

---

```{r}
#| label: session-info
#| code-fold: true
sessionInfo()
```
